### 2019-12-05 paper/paper.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Journal Article
% LaTeX Template
% Version 1.4 (15/5/16)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com) with extensive modifications by
% Vel (vel@LaTeXTemplates.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[twoside,twocolumn]{article}

\usepackage{amsmath}
\usepackage{blindtext} % Package to generate dummy text throughout this template

\usepackage{graphicx}

\usepackage[sc]{mathpazo} % Use the Palatino font
\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\linespread{1.05} % Line spacing - Palatino needs more space between lines
\usepackage{microtype} % Slightly tweak font spacing for aesthetics

\usepackage{float}
\usepackage{cite}

\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\tikzstyle{block} = [rectangle, line width=0.4mm, draw, text width=4em, text centered, minimum height=3em]
\tikzstyle{cloud} = [draw, line width=0.4mm, ellipse, minimum width=10em, node distance=3cm, minimum height=2.5em]
\tikzstyle{line} = [draw, -latex']

\usepackage[english]{babel} % Language hyphenation and typographical rules

\usepackage[hmarginratio=1:1,top=25mm,columnsep=20pt,margin=0.75in]{geometry} % Document margins
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables

\usepackage{subcaption}

\usepackage{lettrine} % The lettrine is the first enlarged letter at the beginning of the text

\usepackage{enumitem} % Customized lists
\setlist[itemize]{noitemsep} % Make itemize lists more compact

\usepackage{abstract} % Allows abstract customization
\renewcommand{\abstractnamefont}{\normalfont\bfseries} % Set the "Abstract" text to bold
\renewcommand{\abstracttextfont}{\normalfont\small\itshape} % Set the abstract itself to small italic text

\usepackage{titlesec} % Allows customization of titles
\renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
\renewcommand\thesubsection{\roman{subsection}} % roman numerals for subsections
\titleformat{\section}[block]{\large\scshape\centering}{\thesection.}{1em}{} % Change the look of the section titles
\titleformat{\subsection}[block]{\large}{\thesubsection.}{1em}{} % Change the look of the section titles

\usepackage{fancyhdr} % Headers and footers
\pagestyle{fancy} % All pages have headers and footers
\fancyhead{} % Blank out the default header
\fancyfoot{} % Blank out the default footer
\fancyhead[C]{Modeling an Economy with Policy Gradients $\bullet$ Autumn 2019 $\bullet$ CS 238} % Custom header text
\fancyfoot[RO,LE]{\thepage} % Custom footer text

\usepackage{titling} % Customizing the title section

\usepackage{hyperref} % For hyperlinks in the PDF

\usepackage{pgfplots}
\usetikzlibrary{intersections}
\usetikzlibrary{patterns}

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\setlength\parindent{0cm}

\setlength{\droptitle}{-4\baselineskip} % Move the title up

\pretitle{\begin{center}\Large\bfseries} % Article title formatting
\posttitle{\end{center}} % Article title closing formatting
\title{Modeling an Economy with Policy Gradients} % Article title
\author{%
\textsc{Nicholas Hirning} \\[1ex]
\normalsize Stanford University \\ 
\normalsize \href{mailto:nhirning@stanford.edu}{nhirning@stanford.edu}
\and
\textsc{Rory Lipkis} \\[1ex]
\normalsize Stanford University \\ 
\normalsize \href{mailto:rlipkis@stanford.edu}{rlipkis@stanford.edu}
\and
\textsc{Andrew Chen} \\[1ex]
\normalsize Stanford University \\ 
\normalsize \href{mailto:asjchen@stanford.edu}{asjchen@stanford.edu}
}
\date{} % Leave empty to omit a date
\renewcommand{\maketitlehookd}{%
\begin{abstract}

TODO

\end{abstract}
}

%----------------------------------------------------------------------------------------

\begin{document}

% Print the title
\maketitle

\section{Introduction}

In simple models of market economies, a common assumption is that each citizen will act in their own best interest. This selfish behavior plays a major role in determining the efficiency of such markets and leads to many of the phenomena predicted and analyzed by economists. The advantage of this assumption is that once a utility function is chosen for these models, each agent in the economy acts simply to maximize their future utility. In practice, this lends itself naturally to construction via a multi-agent reinforcement learning model.

\medskip

A simple model taught in introductory economics classes is the circular flow model. In this setup, there are two types of agents in the economy: firms (or businesses), and people (or households). These agents interact in two different markets, the goods market and the labor market. In the goods market, people pay currency to firms in exchange for goods. In the labor market, firms pay people for their labor to produce more goods. This basic model is diagrammed in figure \ref{fig:circularflow}. The name comes from the fact that currency flows clockwise in a circle between firms and people.

\begin{figure}[h]
  \centering
  \begin{tikzpicture}[node distance = 2cm, auto]
    \node [block] (firms) {firms};
    \node [cloud, below right of=firms] (goods) [below right=-1.4cm and -1cm of firms] {goods market};
    \node [cloud, above right of=firms] (labor) [above right=-1.4cm and -1cm of firms] {labor market};
    \node [block, right of=firms] (people) [right=2.6cm of firms] {people};
    \path [line width=0.4mm, line] (firms) -- (goods);
    \path [line width=0.4mm, line] (firms) -- (labor);
    \path [line width=0.4mm, line] (people) -- (goods);
    \path [line width=0.4mm, line] (people) -- (labor);
  \end{tikzpicture}
  \caption{Basic schematic of the circular flow model, a simple economic model in which society is composed of two types of agents (firms and people) that interact in two distinct markets.} \label{fig:circularflow}
\end{figure}

While the circular flow model does not include many important economic factors, it has the benefit of being simple enough to serve as a starting point for high-level analysis. For more in-depth economic phenomena, one would likely benefit from including more markets and agents; for example, the five-sector economy model includes government, financial markets, and overseas influences. However, as we aim to provide a proof-of-concept for modeling economies with multi-agent reinforcement learning, the circular flow model serves as a simple foundation that can potentially produce nontrivial dynamics.

\medskip

In the following work, we explore how various economic phenomena can be simulated and analyzed by applying reinforcement learning to the circular flow model. While we are limited by the simplicity of the model, there are many economic metrics that can be directly investigated in such a model. For example, the gross domestic product (GDP) is widely touted as a comprehensive measure of a country's economic health. The GDP is defined by the following equation:
\[
  \text{GDP} = C + G + I + N,
\]
where $C$ stands for consumer spending, $G$ for government spending, $I$ for private domestic investments, and $N$ for net exports. In practice, the vast majority of GDP is contained in consumer spending (accounting for roughly $2/3$ of the U.S. GDP). While the circular flow model cannot account for net exports or government spending, the total consumer spending and private domestic investments can be approximated by the total quantity of money exchanged in the goods and labor markets. Thus, such a model can lead to investigation of how GDP changes over time and as a function of the number of agents of each type.

\medskip

Another important economic focus is on income and wealth inequality, both of which are manifested in the circular flow model. A common approach to measuring inequality is to use the Gini coefficient. The Gini coefficient of a distribution $p$ with mean $\mu$ is given by 
\[
  G = \frac{1}{2\mu} \iint p(x)p(y)|x - y|\, \mathrm{d}x\, \mathrm{d}y.
\] 
In a discrete scenario, this just becomes the average absolute difference of all pairs of items (e.g., income or absolute wealth) divided by twice the mean. Given its relevance to modern society, understanding how income inequality develops and progresses over time and as a function of initial wealth and skill distribution is quite important. We aim to perform basic analyses of both of these phenomena in the work presented herein.

\section{Literature Review}

Despite the applicability of reinforcement learning techniques to economics, there are relatively few examples of such applications in existing literature. One particularly relevant example is a paper by Lozano et. al. \cite{RePEc:col:000092:003907} which focuses on modeling an economy ``based on conventions.'' In this work, the authors apply the $\text{SARSA}(\lambda)$ algorithm in a government agent to search for a good expenditure policy. In this case, the authors avoid applying complicated multi-agent reinforcement learning by modeling each firm with a simple algorithm in which firms can copy neighboring firms or (with some low probability) switch to a random policy. The government attempts to increase national income and decrease debt based on some limited knowledge of the existing firms and historical records of income, debt, etc.

\medskip

One important issue that the authors of \cite{RePEc:col:000092:003907} grappled with was the vast (and continuous) state space. In this case, they applied a multilayer-perceptron (MLP) with a single hidden layer to approximate the $Q$-function for $\text{SARSA}(\lambda)$. Given the vast array of actions and states within any economy, approximation with an MLP model or similar is common in such scenarios. The authors found limited success with their model, with the economy collapsing less frequently over time due to the government's policies.

\medskip

Another relevant work by Tesauro \& Kephart involved investigation of the behavior of reinforcement learning agents in the presence of other adaptive agents for economic purposes \cite{pricing-multi-agent}. In particular, Tesauro \& Kephart consider two selling agents competing in an economic model where each agent applied Q-learning. In this case, the state and action space was small enough that the Q-learning could proceed directly. Furthermore, the consumers in the model act via a simple, greedy rule as opposed to acting as an independent, market-shaping force. Another difference between Tesauro \& Kephart and the model proposed here is that the former model does not assume simultaneous price-setting; instead, the two selling agents take turns adjusting prices in response to each other. This turn-based approach is slightly less realistic than the simultaneous setting, but it allowed the authors to frame the problem as a two-player, alternating-turn, arbitrary-sum Markov game. Additionally, the authors spent significant amount of their time computing the Q-function for one of the sellers when the other seller was assumed myopic; they investigate into simultaneous Q-learning as well, but note that no convergence proofs exist for such cases.

\medskip

TODO: Moar.

\section{Methods}

\section{Results and Data Analysis}

\section{Conclusion}

\section{Acknowledgments}

The authors feel that they all contributed equally to the work presented. In particular, Andrew Chen implemented the actor-critic reinforcement learning algorithm, Nicholas Hirning implemented the circular flow model architecture, and Rory Lipkis performed hyperparameter optimization and data analysis. All three authors collaborated on implementing and debugging the REINFORCE policy gradient algorithm and other areas of the code.

\medskip

The authors would like to thank the CS 238 staff for a really great time throughout this course \texttt{;)}.

%----------------------------------------------------------------------------------------
%	REFERENCE LIST
%----------------------------------------------------------------------------------------

\bibliographystyle{unsrt}
\bibliography{sources}

%----------------------------------------------------------------------------------------

\end{document}

### 2019-12-05 firm.py
hyperparameter simplification, linear demand, convergence analysis

@@ -27,7 +27,13 @@ class FirmAgent(Agent):
             units = POSSIBLE_UNITS_FIRM[indices[0]]
             price = POSSIBLE_PRICES_FIRM[indices[1]]
             demand_curve = reciprocal(POSSIBLE_RECIP_DEMAND_PARAMS_FIRM[indices[2]])(NUM_GOODS_MAX_PRODUCE)
-            return Action(price, units, demand_curve)
+        elif self.demand_curve_shape == DemandCurveShape.LINEAR:
+            indices = np.unravel_index(action_num, (POSSIBLE_UNITS_FIRM.shape[0], 
+                POSSIBLE_PRICES_FIRM.shape[0], POSSIBLE_LIN_DEMAND_PARAMS_FIRM.shape[0]))
+            units = POSSIBLE_UNITS_FIRM[indices[0]]
+            price = POSSIBLE_PRICES_FIRM[indices[1]]
+            demand_curve = linear(POSSIBLE_LIN_DEMAND_PARAMS_FIRM[indices[2]])(NUM_GOODS_MAX_PRODUCE)
+        return Action(price, units, demand_curve)
 
     def get_action(self, model):
         if self.rltype == RLType.TRIVIAL:

### 2019-12-05 main.py

@@ -21,11 +21,22 @@ def reinforce_test():
     m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE)
     m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE)
     num_iters = 100
+    avg_GDP = []
     for i in tqdm(range(num_iters)):
-        m.run_episode(100, verbose=False)
+        stats = m.run_episode(100)
+        avg_GDP.append(np.mean(stats['GDP_over_time']))
 
-    stats = m.run_episode(1000, verbose=False)
-    plot_wealth_histories(m, stats)
+    stats = m.run_episode(1000)
+
+    # debugging
+    people_goods_over_time = stats['people_goods_over_time']
+    for i in range(len(people_goods_over_time)):
+        print("skill \t%.3f" % m.people[i].skill)
+        print("goods \t%d" % people_goods_over_time[i][-1])
+        print("money \t%.3f" % m.people[i].money)
+        print("===")
+
+    # plot_wealth_histories(m, stats)
 
 def main():
     reinforce_test()

### 2019-12-05 constants.py
Added actor-critic integration to people and firms, untested

@@ -18,7 +18,7 @@ DISCOUNT = 0.9
 
 class RLType(Enum):
     TRIVIAL = 1
-    DEEPQ = 2
+    Q_ACTOR_CRITIC = 2
     REINFORCE = 3
 
 class DemandCurveShape(Enum):

### 2019-12-05 firm.py

@@ -1,6 +1,7 @@
 import numpy as np 
 import math
 from policy_grad import ReinforcePolicyGradient
+from q_actor_critic import ActorCritic
 from agent import Agent, Action
 from constants import *
 
@@ -12,11 +13,13 @@ class FirmAgent(Agent):
         self.money_recv = []
         self.money_paid = []
 
-        if self.rltype == RLType.REINFORCE:
-            state_dim = 2 * NUM_FIRMS + NUM_PEOPLE 
-            action_dim = POSSIBLE_UNITS_FIRM.shape[0] * POSSIBLE_PRICES_FIRM.shape[0] * POSSIBLE_RECIP_DEMAND_PARAMS_FIRM.shape[0] 
-            self.policy_net = ReinforcePolicyGradient(state_dim, action_dim)
+        state_dim = 2 * NUM_FIRMS + NUM_PEOPLE 
+        action_dim = POSSIBLE_UNITS_FIRM.shape[0] * POSSIBLE_PRICES_FIRM.shape[0] * POSSIBLE_RECIP_DEMAND_PARAMS_FIRM.shape[0] 
 
+        if self.rltype == RLType.REINFORCE:
+            self.policy_net = ReinforcePolicyGradient(state_dim, action_dim)
+        elif self.rltype == RLType.Q_ACTOR_CRITIC:
+            self.actor_critic = ActorCritic(state_dim, action_dim)
 
     # Given a categorical number,
     # returns an Action object for the agents to use
@@ -45,6 +48,11 @@ class FirmAgent(Agent):
             state_input = self.deconstruct_state(model)
             action_num = self.policy_net.choose_action(state_input)
             return self.construct_action(action_num)
+
+        if self.rltype == RLType.Q_ACTOR_CRITIC:
+            state_input = self.deconstruct_state(model)
+            action_num = self.actor_critic.choose_action(state_input)
+            return self.construct_action(action_num)
         
     def reset(self):
         self.money = self.init_money
@@ -56,6 +64,8 @@ class FirmAgent(Agent):
     def end_episode(self):
         if self.rltype == RLType.REINFORCE:
             self.policy_net.update_policy()
+        elif self.rltype == RLType.Q_ACTOR_CRITIC:
+            self.actor_critic.reset_memory()
 
     def update(self, state, action, result):
         """
@@ -67,6 +77,11 @@ class FirmAgent(Agent):
         self.num_goods += goods_recv - goods_sold
         if self.rltype == RLType.REINFORCE:
[…]

### 2019-12-05 person.py

@@ -15,10 +15,12 @@ class PersonAgent(Agent):
         # Per-turn state.
         self.num_hours_to_work = WORK_HOURS_PER_PERSON
 
-        if self.rltype == RLType.REINFORCE:
-            state_dim = 2 * NUM_FIRMS + NUM_PEOPLE 
+        state_dim = 2 * NUM_FIRMS + NUM_PEOPLE 
             action_dim = POSSIBLE_UNITS_PERSON.shape[0] * POSSIBLE_PRICES_PERSON.shape[0] * POSSIBLE_RECIP_DEMAND_PARAMS_PERSON.shape[0] 
+        if self.rltype == RLType.REINFORCE:
             self.policy_net = ReinforcePolicyGradient(state_dim, action_dim)
+        elif self.rltype == RLType.Q_ACTOR_CRITIC:
+            self.actor_critic = ActorCritic(state_dim, action_dim)
 
     # Given a categorical number,
     # returns an Action object for the agents to use
@@ -48,6 +50,11 @@ class PersonAgent(Agent):
             action_num = self.policy_net.choose_action(state_input)
             return self.construct_action(action_num)
 
+        if self.rltype == RLType.Q_ACTOR_CRITIC:
+            state_input = self.deconstruct_state(model)
+            action_num = self.actor_critic.choose_action(state_input)
+            return self.construct_action(action_num)
+
     def reset(self):
         self.money = self.init_money
         self.goods_recv = []
@@ -57,6 +64,8 @@ class PersonAgent(Agent):
     def end_episode(self):
         if self.rltype == RLType.REINFORCE:
             self.policy_net.update_policy()
+        elif self.rltype == RLType.Q_ACTOR_CRITIC:
+            self.actor_critic.reset_memory()
 
     def update(self, state, action, result):
         """
@@ -66,6 +75,11 @@ class PersonAgent(Agent):
         self.money += money_recv - money_paid
         if self.rltype == RLType.REINFORCE:
             self.policy_net.record_reward(goods_recv)
+        elif self.rltype == RLType.Q_ACTOR_CRITIC:
+            self.actor_critic.record_reward(profit)
+            self.actor_critic.update_policy()
+            self.actor_critic.update_q()
+            self.actor_critic.shift_results()
         assert(self.money >= 0)
 
         self.epis_actions.append(action)

### 2019-12-05 q_actor_critic.py

@@ -13,7 +13,7 @@ from constants import *
 # Pseudocode from here: https://towardsdatascience.com/understanding-actor-critic-methods-931b97b6df3f
 class PolicyNetwork(nn.Module):
     def __init__(self, state_dim, action_dim, learning_rate=0.01):
-        super(ReinforcePolicyGradient, self).__init__()
+        super(PolicyNetwork, self).__init__()
         # Should be (2 * number of firms) + (number of people)
         self.state_dim = state_dim 
         # Should be (choices of hours/goods produced) * (choices of prices) * (demand curve parameters)
@@ -37,7 +37,7 @@ class PolicyNetwork(nn.Module):
 
 class QNetwork(nn.Module):
     def __init__(self, state_dim, action_dim, learning_rate=0.01):
-        super(ReinforcePolicyGradient, self).__init__()
+        super(QNetwork, self).__init__()
         # Should be (2 * number of firms) + (number of people)
         self.state_dim = state_dim 
         # Should be (choices of hours/goods produced) * (choices of prices) * (demand curve parameters)
@@ -63,8 +63,8 @@ class ActorCritic(object):
         self.policy_net = PolicyNetwork(state_dim, action_dim)
         self.q_net = QNetwork(state_dim, action_dim)
 
-        self.policy_optimizer = optim.Adam(self.policy_net.parameters())
-        self.q_optimizer = optim.Adam(self.q_net.parameters())
+        self.policy_optimizer = optim.Adam(self.policy_net.parameters(), lr=self.policy_net.learning_rate)
+        self.q_optimizer = optim.Adam(self.q_net.parameters(), lr=self.q_net.learning_rate)
 
         self.gamma = DISCOUNT
 
@@ -79,8 +79,8 @@ class ActorCritic(object):
 
 
     def choose_action(self, state):
-        self.next_state = state 
         state = torch.from_numpy(state).type(torch.FloatTensor)
+        self.next_state = state 
         log_probs = self.policy_net(state)
         probs = torch.exp(log_probs)
         c = Categorical(probs)
@@ -89,7 +89,7 @@ class ActorCritic(object):
         return self.next_action
 
     def update_policy(self):
-        if self.curr_state == None:
+        if type(self.curr_state) == type(None):
             return
         q_value = torch.index_select(self.q_net(self.curr_state), 0, self.curr_action)
         policy_loss = torch.mul(self.curr_action_log_prob, q_value).mul(-1)
[…]

### 2019-12-05 model.py
print stats

@@ -45,13 +45,15 @@ class Model:
         firm_money_recv = [s.money_recv for s in self.firms]
         person_goods_recv = [s.goods_recv for s in self.people]
         person_hours_worked = [s.hours_worked for s in self.people]
+        person_money_hist = [s.money_hist for s in self.people]
         stats = compute_stats(
             self,
             firm_action_hist,
             person_action_hist,
             firm_money_recv,
             firm_money_paid,
-            person_goods_recv
+            person_goods_recv,
+            person_money_hist
         )
 
         # End episode and reset agents.
@@ -67,18 +69,10 @@ class Model:
             person.reset()
 
         if verbose:
-            print("Firm price histories:")
-            pp(stats['firm_price_hists'])
-            print("Firm avg. profit:")
-            pp(stats['firm_avg_profit'])
-            print("Firm avg. first entries:")
-            pp(stats['firm_avg_first_entries'])
-            print("Firm std. first entries:")
-            pp(stats['firm_std_first_entries'])
-            print("Firm money over time:")
-            pp(stats['firm_money_over_time'])
-            print("People goods over time:")
-            pp(stats['people_goods_over_time'])
+            for stat in stats:
+                print(stat)
+                pp(stats[stat])
+                print("=======")
 
             # person_data = (np.mean(person_losses), np.std(person_losses))
             # firm_data = (np.mean(firm_losses), np.std(firm_losses))

### 2019-12-05 person.py

@@ -11,6 +11,7 @@ class PersonAgent(Agent):
         self.epis_actions = []
         self.goods_recv = []
         self.hours_worked = []
+        self.money_hist = []
 
         # Per-turn state.
         self.num_hours_to_work = WORK_HOURS_PER_PERSON
@@ -53,6 +54,7 @@ class PersonAgent(Agent):
         self.goods_recv = []
         self.epis_actions = []
         self.hours_worked = []
+        self.money_hist = [self.money]
 
     def end_episode(self):
         if self.rltype == RLType.REINFORCE:
@@ -71,6 +73,7 @@ class PersonAgent(Agent):
         self.epis_actions.append(action)
         self.goods_recv.append(goods_recv)
         self.hours_worked.append(hours_worked)
+        self.money_hist.append(self.money)
     
     def get_loss(self):
         if self.rltype == RLType.TRIVIAL:

### 2019-12-05 constants.py
Integrated actor-critic with agents

@@ -18,8 +18,9 @@ DISCOUNT = 0.9
 
 class RLType(Enum):
     TRIVIAL = 1
-    Q_ACTOR_CRITIC = 2
+    DEEPQ = 2
     REINFORCE = 3
+    Q_ACTOR_CRITIC = 4
 
 class DemandCurveShape(Enum):
     CONSTANT = 1

### 2019-12-05 firm.py

@@ -89,8 +89,13 @@ class FirmAgent(Agent):
         self.money_paid.append(money_paid)
     
     def get_loss(self):
-        if len(self.policy_net.loss_hist) == 0: return None
-        return self.policy_net.loss_hist[-1]
+        if self.rltype == RLType.REINFORCE:
+            loss_hist = self.policy_net.loss_hist
+        elif self.rltype == RLType.Q_ACTOR_CRITIC:
+            loss_hist = self.actor_critic.policy_loss_hist
+
+        if len(loss_hist) == 0: return None
+        return loss_hist[-1]
     
 
     

### 2019-12-05 person.py

@@ -16,7 +16,7 @@ class PersonAgent(Agent):
         self.num_hours_to_work = WORK_HOURS_PER_PERSON
 
         state_dim = 2 * NUM_FIRMS + NUM_PEOPLE 
-            action_dim = POSSIBLE_UNITS_PERSON.shape[0] * POSSIBLE_PRICES_PERSON.shape[0] * POSSIBLE_RECIP_DEMAND_PARAMS_PERSON.shape[0] 
+        action_dim = POSSIBLE_UNITS_PERSON.shape[0] * POSSIBLE_PRICES_PERSON.shape[0] * POSSIBLE_RECIP_DEMAND_PARAMS_PERSON.shape[0] 
         if self.rltype == RLType.REINFORCE:
             self.policy_net = ReinforcePolicyGradient(state_dim, action_dim)
         elif self.rltype == RLType.Q_ACTOR_CRITIC:
@@ -89,7 +89,12 @@ class PersonAgent(Agent):
     def get_loss(self):
         if self.rltype == RLType.TRIVIAL:
             return 0
-        if len(self.policy_net.loss_hist) == 0: return None
-        return self.policy_net.loss_hist[-1]
+        if self.rltype == RLType.REINFORCE:
+            loss_hist = self.policy_net.loss_hist
+        elif self.rltype == RLType.Q_ACTOR_CRITIC:
+            loss_hist = self.actor_critic.policy_loss_hist
+
+        if len(loss_hist) == 0: return None
+        return loss_hist[-1]
     
     

### 2019-12-05 q_actor_critic.py

@@ -77,6 +77,8 @@ class ActorCritic(object):
         self.next_action_log_prob = None
         self.next_reward = None
 
+        self.policy_loss_hist = []
+
 
     def choose_action(self, state):
         state = torch.from_numpy(state).type(torch.FloatTensor)
@@ -96,6 +98,8 @@ class ActorCritic(object):
         self.policy_optimizer.zero_grad()
         policy_loss.backward()
         self.policy_optimizer.step()
+
+        self.policy_loss_hist.append(policy_loss.item())
         # TODO: do appropriate storing here
 
     def update_q(self):

### 2019-12-05 constants.py
log utility

@@ -4,13 +4,13 @@ import itertools
 
 # These are constants because I use them to initialize 
 # the policy network's input dimension
-NUM_FIRMS = 1
-NUM_PEOPLE = 5
+NUM_FIRMS = 3
+NUM_PEOPLE = 10
 
 NUM_GOODS_MAX_BUY = 10
 NUM_GOODS_MAX_PRODUCE = NUM_GOODS_MAX_BUY * NUM_PEOPLE
 
-WORK_HOURS_PER_PERSON = 40
+WORK_HOURS_PER_PERSON = 10
 TOTAL_MONEY_FIRMS = 10000 * NUM_FIRMS
 TOTAL_MONEY_PEOPLE = 100 * NUM_PEOPLE
 

### 2019-12-05 firm.py

@@ -63,10 +63,12 @@ class FirmAgent(Agent):
         """
         money_paid, money_recv, goods_recv, goods_sold = result
         profit = money_recv - money_paid
+        eps = 0.01
+        utility = np.log(1 + profit / (self.money + eps))
         self.money += profit
         self.num_goods += goods_recv - goods_sold
         if self.rltype == RLType.REINFORCE:
-            self.policy_net.record_reward(profit)
+            self.policy_net.record_reward(utility)
         assert(self.num_goods >= 0 and self.money >= 0)
 
         self.money_recv.append(money_recv)

### 2019-12-05 main.py

@@ -27,7 +27,7 @@ def reinforce_test():
         avg_GDP.append(np.mean(stats['GDP_over_time']))
 
     stats = m.run_episode(1000)
-    print_stats(m, stats)
+    # print_stats(m, stats)
     plot_wealth_histories(m, stats)
 
 def main():

### 2019-12-05 person.py

@@ -7,7 +7,6 @@ class PersonAgent(Agent):
     def __init__(self, money, skill, rltype=RLType.DEEPQ, demand_curve_shape=DemandCurveShape.RECIPROCAL):
         super().__init__(money, rltype, demand_curve_shape)
         self.skill = skill # goods per hour
-
         self.epis_actions = []
         self.goods_recv = []
         self.hours_worked = []
@@ -66,8 +65,10 @@ class PersonAgent(Agent):
         """
         money_paid, money_recv, goods_recv, hours_worked = result
         self.money += money_recv - money_paid
+        eps = 0.01
+        utility = np.log(1 + goods_recv)
         if self.rltype == RLType.REINFORCE:
-            self.policy_net.record_reward(goods_recv)
+            self.policy_net.record_reward(utility)
         assert(self.money >= 0)
 
         self.epis_actions.append(action)

### 2019-12-05 firm.py
Added tracking losses, but currently unstable actor-critic losses

@@ -92,7 +92,7 @@ class FirmAgent(Agent):
         if self.rltype == RLType.REINFORCE:
             loss_hist = self.policy_net.loss_hist
         elif self.rltype == RLType.Q_ACTOR_CRITIC:
-            loss_hist = self.actor_critic.policy_loss_hist
+            loss_hist = list(zip(self.actor_critic.policy_loss_hist, self.actor_critic.q_loss_hist))
 
         if len(loss_hist) == 0: return None
         return loss_hist[-1]

### 2019-12-05 person.py

@@ -95,7 +95,8 @@ class PersonAgent(Agent):
         if self.rltype == RLType.REINFORCE:
             loss_hist = self.policy_net.loss_hist
         elif self.rltype == RLType.Q_ACTOR_CRITIC:
-            loss_hist = self.actor_critic.policy_loss_hist
+            loss_hist = list(zip(self.actor_critic.policy_loss_hist, self.actor_critic.q_loss_hist))
+
 
         if len(loss_hist) == 0: return None
         return loss_hist[-1]

### 2019-12-05 paper/paper.tex
Begin data collection

@@
 Given its relevance to modern society, understanding how income inequality develops and progresses over time and as a function of initial wealth and skill distribution is quite important.
-We aim to perform basic analyses of both of these phenomena in the work presented herein.
+
+\medskip
+
+We aim to explore the applicability of multi-agent reinforcement learning as a means for analyzing macroeconomic phenomena.
+To this end, we implement the circular flow model and perform basic analyses of both GDP and income inequality.
+As this is a multi-agent problem, agents choose actions and learn simultaneously, making the simulation dynamics quite complex.
+Thus, we will also discuss and analyze the effectiveness of two different learning algorithms.
 
@@
 Despite the applicability of reinforcement learning techniques to economics, there are relatively few examples of such applications in existing literature.
-One particularly relevant example is a paper by Lozano et.
-al.
+One particularly relevant example is a paper by Lozano et al.
 \cite{RePEc:col:000092:003907} which focuses on modeling an economy ``based on conventions.'' In this work, the authors apply the $\text{SARSA}(\lambda)$ algorithm in a government agent to search for a good expenditure policy.
@@
 In this case, the state and action space was small enough that the Q-learning could proceed directly.
-Furthermore, the consumers in the model act via a simple, greedy rule as opposed to acting as an independent, market-shaping force.
-Another difference between Tesauro \& Kephart and the model proposed here is that the former model does not assume simultaneous price-setting; instead, the two selling agents take turns adjusting prices in response to each other.
+Furthermore, the consumers in the model acted via a simple, greedy rule as opposed to acting as an independent, market-shaping force.
+Another difference between Tesauro \& Kephart and the model proposed here is that the former model does not assume simultaneous price-setting; instead, the two selling agents took turns adjusting prices in response to each other.
 This turn-based approach is slightly less realistic than the simultaneous setting, but it allowed the authors to frame the problem as a two-player, alternating-turn, arbitrary-sum Markov game.
-Additionally, the authors spent significant amount of their time computing the Q-function for one of the sellers when the other seller was assumed myopic; they investigate into simultaneous Q-learning as well, but note that no convergence proofs exist for such cases.
+Additionally, the authors spent significant amount of their time computing the Q-function for one of the sellers when the other seller was assumed myopic; they investigated simultaneous Q-learning as well, but noted that no convergence proofs exist for such cases.
 
@@
 
-TODO: Moar.
+While there are relatively few examples of reinforcement learning applied to macroeconomic models, there are quite a few papers with relevant material on applying multi-agent models to various settings involving competition and collaboration.
+For example, a 2013 study by Bos et.
+al.
+found that reinforcement learning alone does not accurately model humans in competitive social environments, but it \textit{does} accurately model this when combined with biased reward representations \cite{vandenBos2137}.
+In a separate work, Bereby-Meyer and Roth experiment with reinforcement learning in game theoretic situations, such as the repeated prisoner's dilemma, when complicated by noise \cite{10.1257/aer.96.4.1029}.
+The authors conclude that noise slows learning and strongly negatively affects cooperative tendencies \cite{10.1257/aer.96.4.1029}.
+As such, even with simple multi-agent situations involving some randomness, strong cooperation is unlikely.
+
+\medskip
+
+One study performed by K\"on\"onen studied the situation of two competing brokers selling identical products under two different learning methods: modified gradient descent on the values and a policy gradient approach \cite{doi:10.1002/int.20121}.
+K\"on\"onen found success with both approaches, but found that the system would become intractable as the number of pricing actions increased \cite{doi:10.1002/int.20121}.
+To address this, the author suggested applying an MLP or other generalizable model as an extension.
+In our model, we attempt to combine and extend many of the successful techniques presented in past studies to create an accurate representation of the circular flow model.
 
 \section{Methods}
+
+The circular flow model uses two distinct types of agents: firms and people.
+Correspondingly, the simulation defines these two agent types separately.
+Each agent (firm or human) keeps the following internal state:
+\begin{itemize}
+  \item \textbf{Money.} Each agent can accumulate and spend money during each timestep.
+  \item \textbf{Goods.} Each agent can attain goods (firms create goods through the labor market, and people buy goods through the goods market).
+\end{itemize}
+Each firm receives a reward at each step equal to the profit they generate.
+Thus, firms aim to maximize their future profit.
+Human agents receive a reward equal to the number of goods they consume.
+Furthermore, human agents are each seeded with a constant skill level which governs how many goods a human agent can produce in an hour of work.
+
+\medskip
+
+At each timestep, every agent produces an action that determines how they behave in the two markets.
+In each market, the designated selling agents give a fixed price and number of units.
+The buying agents provide a vectorized demand curve which specifies how much the agent is willing to pay for each subsequent item.
+Thus, each agent provides an action with the following structure:
+\begin{itemize}
+  \item \textbf{Offered Price.} In the labor market, the human agents set their hourly rate. In the goods market, the firms set their price per good.
+  \item \textbf{Offered Units.} In the labor market, the human agents set the number of hours they are willing to work. In the goods market, the firms set the number of goods they are willing to sell.
+  \item \textbf{Demand Curve.} The demand curve is a vector.
+In the labor market, the $i$th entry of the demand curve corresponds to the price the firm will pay to produce its $i$th good.
+In the goods market, the $i$th entry of the demand curve corresponds to the price a human will pay for its $i$th good.
+\end{itemize}
+
+In practice, this action space is vast.
+In order to make the problem tractable, the offered price and offered units are discretized into one of 20 or so values.
+The demand curve is parameterized by TODO.
 

### 2019-12-05 analytics.py

@@ -5,8 +5,11 @@ from constants import *
 from collections import Counter
 
 def compute_stats(m, firm_action_hist, person_action_hist, 
-        firm_money_recv, firm_money_paid, person_goods_recv, person_money_hist):
+        firm_money_recv, firm_money_paid, person_goods_recv, person_money_hist,
+        firm_money_hist):
         stats = {}
+
+        stats['firm_money_hists'] = firm_money_hist
 
         # Frequency of prices offered by firms.
         price_hists = []
@@ -81,6 +84,64 @@ def compute_stats(m, firm_action_hist, person_action_hist,
         stats['gini_over_time'] = list(gini_over_time)
 
         return stats
+
+def save_plots_from_iteration(stats, iteration_num, name):
+    _, axs = plt.subplots(3, 2, figsize=(10, 12))
+    plot_firm_money_hist(axs[0, 0], stats['firm_money_hists'])
+    plot_human_money_hist(axs[0, 1], stats['people_money_over_time'])
+    plot_human_goods_hist(axs[1, 1], stats['people_goods_over_time'])
+    plot_gdp_smoothed(axs[1, 0], stats)
+    plot_median_human_goods(axs[2, 1], stats)
+    plot_gini_coef(axs[2, 0], stats)
+    plt.savefig('%s-iteration-%s' % (name, iteration_num), dpi=300)
+
+def plot_firm_money_hist(ax, firm_money_hists):
+    for idx, firm_money_hist in enumerate(firm_money_hists):
+        ax.plot(np.arange(len(firm_money_hist)), firm_money_hist, label='Firm %s' % idx)
+    ax.set_xlabel("Iteration")
+    ax.set_ylabel("Firm Money")
+    ax.legend()
+
+def plot_human_money_hist(ax, human_money_hists):
+    for idx, human_money_hist in enumerate(human_money_hists):
+        ax.plot(np.arange(len(human_money_hist)), human_money_hist, label='Human %s' % idx)
+    ax.set_xlabel("Iteration")
+    ax.set_ylabel("Human Money")
+    ax.legend()
+
+def plot_human_goods_hist(ax, human_good_hists):
+    for idx, human_good_hist in enumerate(human_good_hists):
+        ax.plot(np.arange(len(human_good_hist)), human_good_hist, label='Human %s' % idx)
+    ax.set_xlabel("Iteration")
[…]

### 2019-12-05 firm.py

@@ -11,6 +11,7 @@ class FirmAgent(Agent):
         self.num_goods = 0
         self.epis_actions = []
         self.money_recv = []
+        self.money_hist = [self.money]
         self.money_paid = []
 
         state_dim = 2 * NUM_FIRMS + NUM_PEOPLE 
@@ -60,6 +61,7 @@ class FirmAgent(Agent):
         self.epis_actions = []
         self.money_recv = []
         self.money_paid = []
+        self.money_hist = [self.init_money]
 
     def end_episode(self):
         if self.rltype == RLType.REINFORCE:
@@ -89,6 +91,7 @@ class FirmAgent(Agent):
         self.money_recv.append(money_recv)
         self.epis_actions.append(action)
         self.money_paid.append(money_paid)
+        self.money_hist.append(self.money)
     
     def get_loss(self):
         if self.rltype == RLType.REINFORCE:

### 2019-12-05 main.py

@@ -18,15 +18,17 @@ def trivial_test():
 
 def reinforce_test():
     m = model.Model(10000)
-    m.create_firms(NUM_FIRMS, rltype=RLType.Q_ACTOR_CRITIC)
+    m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE)
     m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE)
-    num_iters = 100
+    num_iters = 75
     avg_GDP = []
-    for i in tqdm(range(num_iters)):
+    for i in range(num_iters):
         stats = m.run_episode(100, verbose=True)
-        avg_GDP.append(np.mean(stats['GDP_over_time']))
+        save_plots_from_iteration(stats, i, 'plots/test4-nolog/test')
+        # avg_GDP.append(np.mean(stats['GDP_over_time']))
 
-    stats = m.run_episode(1000)
+    stats = m.run_episode(500)
+    save_plots_from_iteration(stats, i, 'plots/test4-nolog/test-final')
     # print_stats(m, stats)
     plot_wealth_histories(m, stats)
 

### 2019-12-05 person.py

@@ -75,7 +75,7 @@ class PersonAgent(Agent):
         money_paid, money_recv, goods_recv, hours_worked = result
         self.money += money_recv - money_paid
         eps = 0.01
-        utility = np.log(1 + goods_recv)
+        utility = goods_recv #np.log(1 + goods_recv)
         if self.rltype == RLType.REINFORCE:
             self.policy_net.record_reward(utility)
         elif self.rltype == RLType.Q_ACTOR_CRITIC:
@@ -84,6 +84,8 @@ class PersonAgent(Agent):
             self.actor_critic.update_q()
             self.actor_critic.shift_results()
         assert(self.money >= 0)
+
+        self.money *= 1.01
 
         self.epis_actions.append(action)
         self.goods_recv.append(goods_recv)

### 2019-12-05 paper/paper.tex
Starting RL section of paper

--- 2019-12-05 paper/paper.tex
@@
 \section{Methods}
+\subsection{Problem Framing******}
+
+[multiagent -> individual RL agents]
+[describe state]
+[describe actions]
+[describe rewards]
+[parameterizing the demand curve]
+[full transparency]
+
+\subsection{RL Algorithms******}
+We worked within the realm of policy gradient methods; we chose this family of algorithms
+
+[exploration strategy is built in?]
+
+with large action spaces.
+
+\subsubsection{REINFORCE Policy Gradient******}
 
@@
 The authors feel that they all contributed equally to the work presented.
-In particular, Andrew Chen implemented the actor-critic reinforcement learning algorithm, Nicholas Hirning implemented the circular flow model architecture, and Rory Lipkis performed hyperparameter optimization and data analysis.
+In particular, Andrew Chen implemented the policy gradient and actor-critic algorithms, Nicholas Hirning implemented the circular flow model architecture, and Rory Lipkis performed hyperparameter optimization and data analysis.
 All three authors collaborated on implementing and debugging the REINFORCE policy gradient algorithm and other areas of the code.

### 2019-12-05 paper/paper.tex
Fixed methods section

--- 2019-12-05 paper/paper.tex
@@
 
+[multiagent -> individual RL agents]
+[describe state, $s_s$]
+[describe actions, $s_a$, as discretized to more easily track the]
+[describe rewards]
+[parameterizing the demand curve]
+[full transparency of staters]
+
+\subsection{RL Algorithms******}
+We worked within the realm of policy gradient methods; we chose this family of algorithms
+
+[exploration strategy is built in?]
+
+with large action spaces.
+
+As framed, our problem contains large — in fact, high dimensional — state spaces.
+Specifically, the dimension of a given state is $s_s = 2 \cdot (\text{number of firms}) + (\text{number of people})$.
+Furthermore, while the values in each dimension were integers, they could take on a wide range of values.
+Therefore, in every policy, we use neural network architectures to convert a given state to a vector, containing log probabilities of taking each action.
+In empirical testing, if $s_a$ is the number of possible actions, we've seen that the following architecture is a simple framework that still yields convergence in our context:
+\begin{itemize}
+  \item Input layer with state input, of dimension $s_s$. State values are unnormalized.
+  \item Hidden layer with tanh activation, of dimension $\left\lfloor \sqrt{s_s \, s_a} \right\rfloor$
+  \item Output layer with tanh activation, followed by log softmax, of dimension $s_a$
+\end{itemize}
+During training, we generally use a learning rate of $\alpha = 0.01$.
+
+\subsubsection{REINFORCE Policy Gradient******}
+
 \section{Results and Data Analysis}

### 2019-12-05 analytics.py
Updates to analytics

@@ -1,6 +1,7 @@
 import model
 import numpy as np
 import matplotlib.pyplot as plt
+from matplotlib import cm
 from constants import *
 from collections import Counter
 
@@ -8,6 +9,8 @@ def compute_stats(m, firm_action_hist, person_action_hist,
         firm_money_recv, firm_money_paid, person_goods_recv, person_money_hist,
         firm_money_hist):
         stats = {}
+
+        stats['person_skills'] = [p.skill for p in m.people]
 
         stats['firm_money_hists'] = firm_money_hist
 
@@ -86,12 +89,13 @@ def compute_stats(m, firm_action_hist, person_action_hist,
         return stats
 
 def save_plots_from_iteration(stats, iteration_num, name):
-    _, axs = plt.subplots(3, 2, figsize=(10, 12))
+    _, axs = plt.subplots(3, 3, figsize=(15, 12))
     plot_firm_money_hist(axs[0, 0], stats['firm_money_hists'])
-    plot_human_money_hist(axs[0, 1], stats['people_money_over_time'])
+    plot_human_money_hist(axs[0, 1], stats)
     plot_human_goods_hist(axs[1, 1], stats['people_goods_over_time'])
     plot_gdp_smoothed(axs[1, 0], stats)
     plot_median_human_goods(axs[2, 1], stats)
+    plot_total_money(axs[0, 2], stats)
     plot_gini_coef(axs[2, 0], stats)
     plt.savefig('%s-iteration-%s' % (name, iteration_num), dpi=300)
 
@@ -102,12 +106,20 @@ def plot_firm_money_hist(ax, firm_money_hists):
     ax.set_ylabel("Firm Money")
     ax.legend()
 
-def plot_human_money_hist(ax, human_money_hists):
-    for idx, human_money_hist in enumerate(human_money_hists):
-        ax.plot(np.arange(len(human_money_hist)), human_money_hist, label='Human %s' % idx)
+def plot_human_money_hist(ax, stats):
+    colormap = cm.get_cmap('binary', len(stats['people_money_over_time']))
+    skill_map = sorted([(skill, idx) for idx, skill in enumerate(stats['person_skills'])])
+    max_skill = skill_map[-1][0]
+    min_skill = skill_map[0][0]
+    dskill = max_skill - min_skill
+    
+    human_money_hists = stats['people_money_over_time']
+    for skill, idx in skill_map:
+        human_money_hist = human_money_hists[idx]
[…]

### 2019-12-05 constants.py

@@ -4,17 +4,21 @@ import itertools
 
 # These are constants because I use them to initialize 
 # the policy network's input dimension
-NUM_FIRMS = 5
-NUM_PEOPLE = 15
+NUM_FIRMS = 3
+NUM_PEOPLE = 8
 
 NUM_GOODS_MAX_BUY = 10
-NUM_GOODS_MAX_PRODUCE = NUM_GOODS_MAX_BUY * NUM_PEOPLE
-
 WORK_HOURS_PER_PERSON = 10
 TOTAL_MONEY_FIRMS = 10000 * NUM_FIRMS
 TOTAL_MONEY_PEOPLE = 1000 * NUM_PEOPLE
+DISCOUNT = 0.99
+HUMAN_INTEREST = 1.01
+FIRM_OPERATING_COST = 20
 
-DISCOUNT = 0.95
+# Length of firm demand curve.
+NUM_GOODS_MAX_PRODUCE = NUM_GOODS_MAX_BUY * NUM_PEOPLE
+
+
 
 class RLType(Enum):
     TRIVIAL = 1

### 2019-12-05 firm.py

@@ -88,6 +88,8 @@ class FirmAgent(Agent):
             self.actor_critic.shift_results()
         assert(self.num_goods >= 0 and self.money >= 0)
 
+        self.money -= FIRM_OPERATING_COST
+
         self.money_recv.append(money_recv)
         self.epis_actions.append(action)
         self.money_paid.append(money_paid)

### 2019-12-05 main.py

@@ -20,15 +20,16 @@ def reinforce_test():
     m = model.Model(10000)
     m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE)
     m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE)
-    num_iters = 75
+    print("Person Skills: %s" % [round(p.skill, 2) for p in m.people])
+    num_iters = 20
     avg_GDP = []
     for i in range(num_iters):
         stats = m.run_episode(100, verbose=True)
-        save_plots_from_iteration(stats, i, 'plots/test4-nolog/test')
+        save_plots_from_iteration(stats, i, 'plots/test5/test')
         # avg_GDP.append(np.mean(stats['GDP_over_time']))
 
-    stats = m.run_episode(500)
-    save_plots_from_iteration(stats, i, 'plots/test4-nolog/test-final')
+    stats = m.run_episode(100)
+    save_plots_from_iteration(stats, i, 'plots/test5/test-final')
     # print_stats(m, stats)
     plot_wealth_histories(m, stats)
 

### 2019-12-05 model.py

@@ -78,7 +78,6 @@ class Model:
 
             losses = (np.mean(person_losses), np.mean(firm_losses), np.mean(np.concatenate((person_losses, firm_losses))))
             print("Losses (P, F, T) = (%8.5f, %8.5f, %8.5f)" % losses)
-            print("Person Skills: %s" % [round(p.skill, 2) for p in self.people])
 
         return stats
 

### 2019-12-05 person.py

@@ -11,6 +11,7 @@ class PersonAgent(Agent):
         self.goods_recv = []
         self.hours_worked = []
         self.money_hist = []
+        self.iteration = 0
 
         # Per-turn state.
         self.num_hours_to_work = WORK_HOURS_PER_PERSON
@@ -61,6 +62,7 @@ class PersonAgent(Agent):
         self.epis_actions = []
         self.hours_worked = []
         self.money_hist = [self.money]
+        self.iteration = 0
 
     def end_episode(self):
         if self.rltype == RLType.REINFORCE:
@@ -72,9 +74,11 @@ class PersonAgent(Agent):
         """
         @param result - (money paid, money recv, goods recv, hours worked)
         """
+        self.iteration += 1
+
+        # Update internal action function.
         money_paid, money_recv, goods_recv, hours_worked = result
         self.money += money_recv - money_paid
-        eps = 0.01
         utility = goods_recv #np.log(1 + goods_recv)
         if self.rltype == RLType.REINFORCE:
             self.policy_net.record_reward(utility)
@@ -85,7 +89,8 @@ class PersonAgent(Agent):
             self.actor_critic.shift_results()
         assert(self.money >= 0)
 
-        self.money *= 1.01
+        # Interest update.
+        self.money *= HUMAN_INTEREST
 
         self.epis_actions.append(action)
         self.goods_recv.append(goods_recv)

### 2019-12-05 analytics.py
More analytics

@@ -4,6 +4,8 @@ import matplotlib.pyplot as plt
 from matplotlib import cm
 from constants import *
 from collections import Counter
+import seaborn as sns
+sns.set(style="darkgrid")
 
 def compute_stats(m, firm_action_hist, person_action_hist, 
         firm_money_recv, firm_money_paid, person_goods_recv, person_money_hist,
@@ -92,7 +94,7 @@ def save_plots_from_iteration(stats, iteration_num, name):
     _, axs = plt.subplots(3, 3, figsize=(15, 12))
     plot_firm_money_hist(axs[0, 0], stats['firm_money_hists'])
     plot_human_money_hist(axs[0, 1], stats)
-    plot_human_goods_hist(axs[1, 1], stats['people_goods_over_time'])
+    plot_human_goods_hist(axs[1, 1], stats)
     plot_gdp_smoothed(axs[1, 0], stats)
     plot_median_human_goods(axs[2, 1], stats)
     plot_total_money(axs[0, 2], stats)
@@ -107,7 +109,7 @@ def plot_firm_money_hist(ax, firm_money_hists):
     ax.legend()
 
 def plot_human_money_hist(ax, stats):
-    colormap = cm.get_cmap('binary', len(stats['people_money_over_time']))
+    colormap = cm.get_cmap('cividis', len(stats['people_money_over_time']))
     skill_map = sorted([(skill, idx) for idx, skill in enumerate(stats['person_skills'])])
     max_skill = skill_map[-1][0]
     min_skill = skill_map[0][0]
@@ -116,28 +118,44 @@ def plot_human_money_hist(ax, stats):
     human_money_hists = stats['people_money_over_time']
     for skill, idx in skill_map:
         human_money_hist = human_money_hists[idx]
-        color = colormap((skill - min_skill) / dskill)
+        color = colormap((skill - min_skill) / (dskill))
         ax.plot(np.arange(len(human_money_hist)), human_money_hist, color=color)
     ax.set_xlabel("Iteration")
     ax.set_ylabel("Human Money")
 
-def plot_human_goods_hist(ax, human_good_hists):
-    for idx, human_good_hist in enumerate(human_good_hists):
-        ax.plot(np.arange(len(human_good_hist)), human_good_hist, label='Human %s' % idx)
+def plot_human_goods_hist(ax, stats):
+    human_good_hists = stats['people_goods_over_time']
+
+    colormap = cm.get_cmap('cividis', len(stats['people_money_over_time']))
[…]

### 2019-12-05 constants.py

@@ -13,6 +13,7 @@ TOTAL_MONEY_FIRMS = 10000 * NUM_FIRMS
 TOTAL_MONEY_PEOPLE = 1000 * NUM_PEOPLE
 DISCOUNT = 0.99
 HUMAN_INTEREST = 1.01
+FIRM_INTEREST = 0.996
 FIRM_OPERATING_COST = 20
 
 # Length of firm demand curve.

### 2019-12-05 firm.py

@@ -89,6 +89,7 @@ class FirmAgent(Agent):
         assert(self.num_goods >= 0 and self.money >= 0)
 
         self.money -= FIRM_OPERATING_COST
+        self.money *= FIRM_INTEREST
 
         self.money_recv.append(money_recv)
         self.epis_actions.append(action)

### 2019-12-05 main.py
fixed linear

@@ -18,8 +18,8 @@ def trivial_test():
 
 def reinforce_test():
     m = model.Model(10000)
-    m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE)
-    m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE)
+    m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.LINEAR)
+    m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.LINEAR)
     num_iters = 75
     avg_GDP = []
     for i in range(num_iters):

### 2019-12-05 person.py

@@ -16,7 +16,13 @@ class PersonAgent(Agent):
         self.num_hours_to_work = WORK_HOURS_PER_PERSON
 
         state_dim = 2 * NUM_FIRMS + NUM_PEOPLE 
-        action_dim = POSSIBLE_UNITS_PERSON.shape[0] * POSSIBLE_PRICES_PERSON.shape[0] * POSSIBLE_RECIP_DEMAND_PARAMS_PERSON.shape[0] 
+        demand_params = None
+        if self.demand_curve_shape == DemandCurveShape.RECIPROCAL:
+            demand_params = POSSIBLE_RECIP_DEMAND_PARAMS_PERSON
+        elif self.demand_curve_shape == DemandCurveShape.LINEAR:
+            demand_params = POSSIBLE_LIN_DEMAND_PARAMS_PERSON
+
+        action_dim = POSSIBLE_UNITS_PERSON.shape[0] * POSSIBLE_PRICES_PERSON.shape[0] * demand_params.shape[0] 
         if self.rltype == RLType.REINFORCE:
             self.policy_net = ReinforcePolicyGradient(state_dim, action_dim)
         elif self.rltype == RLType.Q_ACTOR_CRITIC:

### 2019-12-06 analytics.py
Push to pull Rorys new plots

@@ -9,12 +9,13 @@ sns.set(style="darkgrid")
 
 def compute_stats(m, firm_action_hist, person_action_hist, 
         firm_money_recv, firm_money_paid, person_goods_recv, person_money_hist,
-        firm_money_hist):
+        firm_money_hist, firm_goods_hist):
         stats = {}
 
         stats['person_skills'] = [p.skill for p in m.people]
 
         stats['firm_money_hists'] = firm_money_hist
+        stats['firm_goods_hists'] = firm_goods_hist
 
         # Frequency of prices offered by firms.
         price_hists = []
@@ -95,10 +96,11 @@ def save_plots_from_iteration(stats, iteration_num, name):
     plot_firm_money_hist(axs[0, 0], stats['firm_money_hists'])
     plot_human_money_hist(axs[0, 1], stats)
     plot_human_goods_hist(axs[1, 1], stats)
-    plot_gdp_smoothed(axs[1, 0], stats)
+    plot_gdp_smoothed(axs[2, 2], stats)
     plot_median_human_goods(axs[2, 1], stats)
     plot_total_money(axs[0, 2], stats)
     plot_gini_coef(axs[2, 0], stats)
+    plot_firm_goods(axs[1, 0], stats)
     plt.savefig('%s-iteration-%s' % (name, iteration_num), dpi=300)
 
 def plot_firm_money_hist(ax, firm_money_hists):
@@ -170,6 +172,13 @@ def plot_total_money(ax, stats):
     ax.set_xlabel("Iteration")
     ax.set_ylabel("Total Money")
 
+def plot_firm_goods(ax, stats):
+    fg = stats['firm_goods_hists']
+    for idx, firm_good_hist in enumerate(fg):
+        ax.plot(np.arange(len(firm_good_hist)), firm_good_hist, label='Firm %s' % idx)
+    ax.set_xlabel("Iteration")
+    ax.set_ylabel("Firm Goods")
+    ax.legend()
 
 def plot_human_wealth(ax, stats):
     #TODO: Rory, combine money + goods ==> wealth 

### 2019-12-06 constants.py

@@ -4,8 +4,8 @@ import itertools
 
 # These are constants because I use them to initialize 
 # the policy network's input dimension
-NUM_FIRMS = 3
-NUM_PEOPLE = 8
+NUM_FIRMS = 5
+NUM_PEOPLE = 25
 
 NUM_GOODS_MAX_BUY = 10
 WORK_HOURS_PER_PERSON = 10
@@ -13,7 +13,7 @@ TOTAL_MONEY_FIRMS = 10000 * NUM_FIRMS
 TOTAL_MONEY_PEOPLE = 1000 * NUM_PEOPLE
 DISCOUNT = 0.99
 HUMAN_INTEREST = 1.01
-FIRM_INTEREST = 0.996
+FIRM_INTEREST = 0.999
 FIRM_OPERATING_COST = 20
 
 # Length of firm demand curve.

### 2019-12-06 firm.py

@@ -13,6 +13,7 @@ class FirmAgent(Agent):
         self.money_recv = []
         self.money_hist = [self.money]
         self.money_paid = []
+        self.goods_hist = [0]
 
         state_dim = 2 * NUM_FIRMS + NUM_PEOPLE 
         action_dim = POSSIBLE_UNITS_FIRM.shape[0] * POSSIBLE_PRICES_FIRM.shape[0] * POSSIBLE_RECIP_DEMAND_PARAMS_FIRM.shape[0] 
@@ -61,6 +62,7 @@ class FirmAgent(Agent):
         self.epis_actions = []
         self.money_recv = []
         self.money_paid = []
+        self.goods_hist = [0]
         self.money_hist = [self.init_money]
 
     def end_episode(self):
@@ -89,8 +91,10 @@ class FirmAgent(Agent):
         assert(self.num_goods >= 0 and self.money >= 0)
 
         self.money -= FIRM_OPERATING_COST
+        self.money = max(0, self.money)
         self.money *= FIRM_INTEREST
 
+        self.goods_hist.append(self.num_goods)
         self.money_recv.append(money_recv)
         self.epis_actions.append(action)
         self.money_paid.append(money_paid)

### 2019-12-06 main.py

@@ -21,17 +21,17 @@ def reinforce_test():
     m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE)
     m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE)
     print("Person Skills: %s" % [round(p.skill, 2) for p in m.people])
-    num_iters = 20
+    num_iters = 100
     avg_GDP = []
     for i in range(num_iters):
-        stats = m.run_episode(100, verbose=True)
-        save_plots_from_iteration(stats, i, 'plots/test5/test')
+        stats = m.run_episode(200, verbose=True)
+        save_plots_from_iteration(stats, i, 'plots/test7/test')
         # avg_GDP.append(np.mean(stats['GDP_over_time']))
 
-    stats = m.run_episode(100)
-    save_plots_from_iteration(stats, i, 'plots/test5/test-final')
+    stats = m.run_episode(200)
+    save_plots_from_iteration(stats, i, 'plots/test7/test-final')
     # print_stats(m, stats)
-    plot_wealth_histories(m, stats)
+    # plot_wealth_histories(m, stats)
 
 def main():
     reinforce_test()

### 2019-12-06 model.py

@@ -47,6 +47,7 @@ class Model:
         person_hours_worked = [s.hours_worked for s in self.people]
         person_money_hist = [s.money_hist for s in self.people]
         firm_money_hist = [s.money_hist for s in self.firms]
+        firm_goods_hist = [s.goods_hist for s in self.firms]
         stats = compute_stats(
             self,
             firm_action_hist,
@@ -55,7 +56,8 @@ class Model:
             firm_money_paid,
             person_goods_recv,
             person_money_hist,
-            firm_money_hist
+            firm_money_hist,
+            firm_goods_hist
         )
 
         # End episode and reset agents.

### 2019-12-06 policy_grad.py

@@ -42,6 +42,7 @@ class ReinforcePolicyGradient(nn.Module):
             self.layer1,
             nn.Tanh(),
             self.layer2,
+            nn.Dropout(0.5)
             # nn.Softmax()
         )
         # return model(x)

### 2019-12-06 analytics.py
market price of goods analysis

@@ -88,18 +88,30 @@ def compute_stats(m, firm_action_hist, person_action_hist,
             gini_over_time /= 2 * n * p_tot
         stats['gini_over_time'] = list(gini_over_time)
 
+        # Market price of goods.
+        goods_tot = 0
+        for p in person_goods_recv:
+            goods_tot += np.array(p)
+        money_tot = 0
+        for f in firm_money_recv:
+            money_tot += np.array(f)
+        stats['market_price_goods'] = list(money_tot / goods_tot)
+
         return stats
 
 def save_plots_from_iteration(stats, iteration_num, name):
     _, axs = plt.subplots(3, 3, figsize=(15, 12))
     plot_firm_money_hist(axs[0, 0], stats['firm_money_hists'])
     plot_human_money_hist(axs[0, 1], stats)
-    plot_human_goods_hist(axs[1, 1], stats)
-    plot_gdp_smoothed(axs[1, 0], stats)
-    plot_median_human_goods(axs[2, 1], stats)
     plot_total_money(axs[0, 2], stats)
+    plot_gdp_smoothed(axs[1, 0], stats)
+    plot_human_goods_hist(axs[1, 1], stats)
+    plot_market_price_goods(axs[1, 2], stats)
     plot_gini_coef(axs[2, 0], stats)
+    plot_median_human_goods(axs[2, 1], stats)
+
     plt.savefig('%s-iteration-%s' % (name, iteration_num), dpi=300)
+    plt.close('all')
 
 def plot_firm_money_hist(ax, firm_money_hists):
     for idx, firm_money_hist in enumerate(firm_money_hists):
@@ -109,7 +121,7 @@ def plot_firm_money_hist(ax, firm_money_hists):
     ax.legend()
 
 def plot_human_money_hist(ax, stats):
-    colormap = cm.get_cmap('cividis', len(stats['people_money_over_time']))
+    colormap = cm.get_cmap('plasma', len(stats['people_money_over_time']))
     skill_map = sorted([(skill, idx) for idx, skill in enumerate(stats['person_skills'])])
     max_skill = skill_map[-1][0]
     min_skill = skill_map[0][0]
@@ -126,7 +138,7 @@ def plot_human_money_hist(ax, stats):
 def plot_human_goods_hist(ax, stats):
     human_good_hists = stats['people_goods_over_time']
 
-    colormap = cm.get_cmap('cividis', len(stats['people_money_over_time']))
[…]

### 2019-12-06 analytics.py
Done testing

@@ -194,7 +194,9 @@ def plot_firm_goods(ax, stats):
     ax.legend()
 
 def plot_market_price_goods(ax, stats):
-    ax.plot(stats['market_price_goods'])
+
+    ax.plot(stats['market_price_goods'], alpha=0.4)
+    ax.plot(smooth(stats['market_price_goods'], k=5))
     ax.set_xlabel("Iteration")
     ax.set_ylabel("Market Price of Goods")
 

### 2019-12-06 constants.py

@@ -13,8 +13,8 @@ TOTAL_MONEY_FIRMS = 10000 * NUM_FIRMS
 TOTAL_MONEY_PEOPLE = 1000 * NUM_PEOPLE
 DISCOUNT = 0.99
 HUMAN_INTEREST = 1.00 + 0.05 / NUM_PEOPLE
-FIRM_INTEREST = 1.00 - 0.02 / NUM_FIRMS
-FIRM_OPERATING_COST = 20
+FIRM_INTEREST = 1.00 #- 0.0075 / NUM_FIRMS
+FIRM_OPERATING_COST = 10 # 20
 
 # Length of firm demand curve.
 NUM_GOODS_MAX_PRODUCE = NUM_GOODS_MAX_BUY * NUM_PEOPLE

### 2019-12-06 main.py

@@ -24,12 +24,12 @@ def reinforce_test():
     num_iters = 100
     avg_GDP = []
     for i in range(num_iters):
-        stats = m.run_episode(200, verbose=True)
-        save_plots_from_iteration(stats, i, 'plots/test8/test')
+        stats = m.run_episode(300, verbose=True)
+        save_plots_from_iteration(stats, i, 'plots/test9/test')
         # avg_GDP.append(np.mean(stats['GDP_over_time']))
 
-    stats = m.run_episode(200)
-    save_plots_from_iteration(stats, i, 'plots/test8/test-final')
+    stats = m.run_episode(300)
+    save_plots_from_iteration(stats, i, 'plots/test9/test-final')
     # print_stats(m, stats)
     # plot_wealth_histories(m, stats)
 

### 2019-12-06 analytics.py
wage analytics

@@ -13,7 +13,6 @@ def compute_stats(m, firm_action_hist, person_action_hist,
         stats = {}
 
         stats['person_skills'] = [p.skill for p in m.people]
-
         stats['firm_money_hists'] = firm_money_hist
         stats['firm_goods_hists'] = firm_goods_hist
 
@@ -98,10 +97,19 @@ def compute_stats(m, firm_action_hist, person_action_hist,
             money_tot += np.array(f)
         stats['market_price_goods'] = list(money_tot / goods_tot)
 
+        # Market price of labor
+        hours_tot = 0
+        for p in m.people:
+            hours_tot += np.array(p.hours_worked)
+        money_tot = 0
+        for f in firm_money_paid:
+            money_tot += np.array(f)
+        stats['market_price_labor'] = list(money_tot / hours_tot)
+
         return stats
 
 def save_plots_from_iteration(stats, iteration_num, name):
-    _, axs = plt.subplots(3, 3, figsize=(15, 12))
+    _, axs = plt.subplots(4, 4, figsize=(15, 12))
     plot_firm_money_hist(axs[0, 0], stats['firm_money_hists'])
     plot_human_money_hist(axs[0, 1], stats)
     plot_human_goods_hist(axs[1, 1], stats)
@@ -113,6 +121,7 @@ def save_plots_from_iteration(stats, iteration_num, name):
     plot_market_price_goods(axs[1, 2], stats)
     plot_gini_coef(axs[2, 0], stats)
     plot_firm_goods(axs[1, 0], stats)
+    plot_market_price_labor(axs[3, 0], stats)
     plt.savefig('%s-iteration-%s' % (name, iteration_num), dpi=300)
     plt.close('all')
 
@@ -197,6 +206,11 @@ def plot_market_price_goods(ax, stats):
     ax.plot(stats['market_price_goods'])
     ax.set_xlabel("Iteration")
     ax.set_ylabel("Market Price of Goods")
+
+def plot_market_price_labor(ax, stats):
+    ax.plot(stats['market_price_labor'])
+    ax.set_xlabel("Iteration")
+    ax.set_ylabel("Market Price of Labor")
 
 def plot_human_wealth(ax, stats):
     #TODO: Rory, combine money + goods ==> wealth 

### 2019-12-06 analytics.py
Working on new stuff

@@ -7,6 +7,9 @@ from collections import Counter
 import seaborn as sns
 sns.set(style="darkgrid")
 
+def create_loss_plot(stats, name):
+    pass
+
 def compute_stats(m, firm_action_hist, person_action_hist, 
         firm_money_recv, firm_money_paid, person_goods_recv, person_money_hist,
         firm_money_hist, firm_goods_hist):

### 2019-12-06 constants.py

@@ -4,8 +4,8 @@ import itertools
 
 # These are constants because I use them to initialize 
 # the policy network's input dimension
-NUM_FIRMS = 5
-NUM_PEOPLE = 25
+NUM_FIRMS = 1
+NUM_PEOPLE = 5
 
 NUM_GOODS_MAX_BUY = 10
 WORK_HOURS_PER_PERSON = 10

### 2019-12-06 main.py

@@ -16,6 +16,15 @@ def trivial_test():
     for person in m.people:
         print(person.money)
 
+def plot1_for_paper():
+    m = model.Model(1e5)
+    m.create_firms(1, rltype=RLType.REINFORCE)
+    m.create_people(5, rltype=RLType.TRIVIAL)
+    num_iters = 100
+    for i in range(num_iters):
+        stats = m.run_episode(200, verbose=True)
+        
+
 def reinforce_test():
     m = model.Model(10000)
     m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.LINEAR)

### 2019-12-06 model.py

@@ -74,6 +74,9 @@ class Model:
             person_losses.append(person.get_loss())
             person.reset()
 
+        stats['firm_losses'] = firm_losses
+        stats['person_losses'] = person_losses
+
         if verbose:
             # for stat in stats:
             #     print(stat)

### 2019-12-06 paper/paper.tex
Corrected paper RL section with logs, pre-advantage AC

@@
 \usepackage{cite}
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
 
@@
 \subsection{RL Algorithms******}
-We worked within the realm of policy gradient methods; we chose this family of algorithms
-
-[exploration strategy is built in?]
-
-with large action spaces.
-
+Following the literature we reviewed, we worked within the realm of policy gradient methods; we also choose this family of algorithms because they have stronger convergence guarantees than value-based RL methods (albeit potentially to local optimum that are not global optimum).
+\newline\newline
 As framed, our problem contains large — in fact, high dimensional — state spaces.
@@
   \item Input layer with state input, of dimension $s_s$. State values are unnormalized.
-  \item Hidden layer with tanh activation, of dimension $\left\lfloor \sqrt{s_s \, s_a} \right\rfloor$
-  \item Output layer with tanh activation, followed by log softmax, of dimension $s_a$
+  \item Fully-connected hidden layer with tanh activation, of dimension $\left\lfloor \sqrt{s_s \, s_a} \right\rfloor$
+  \item Fully-connected output layer with tanh activation, followed by log softmax, of dimension $s_a$
 \end{itemize}
@@
 
-\subsubsection{REINFORCE Policy Gradient******}
+\subsubsection{REINFORCE Policy Gradient}
+Perhaps the simplest policy gradient method is the REINFORCE algorithm.
+This algorithm uses a Monte Carlo-type approach to gathering training data.
+We manually specify the length of the episode, generally using $T = 100$.
+\begin{algorithm}
+\caption{REINFORCE Policy Gradient}\label{euclid}
+\begin{algorithmic}[1]
+\Procedure{REINFORCE}{}
+\State $\text{Initialize policy network parameters }\theta$
+\For {$i = 1, 2, \ldots, M$}
+\State $\text{Follow current policy to obtain trajectory }$\newline\phantom{======}$\{s_1, a_1, r_1, \ldots, s_T, a_T, r_T\}$.
+\For {$t = 1, 2, \ldots, T$}
+\State $\text{Compute }v_t = \sum_{i = t}^n \gamma^{i- t} r_i$.
+\State $\text{Update } \theta \leftarrow \theta + \alpha \nabla_{\theta}\log\pi_{\theta} (s_t, a_t) v_t$
+\EndFor
+\EndFor
+\State \Return {$\theta$}
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+This algorithm runs reasonably quickly per episode. However, we also need to run more episodes because:
+\begin{itemize}
+  \item Agents only update the parameters once per episode.
+  \item In many contexts, vanilla REINFORCE has been known to suffer from high variance in the losses from episode to episode.
+  \item In our multi-agent problem, the environment dynamics constantly change because every agent responds to each timestep in the markets.
+\end{itemize}
+
+\subsubsection{Q Actor-Critic}
+An alternative family of policy gradient methods is the \textit{actor-critic} family.
+In these algorithms, the actor component computes the log probabilities $\log \pi_{\theta}(s, a)$ while the critic component approximates the functions $Q(s, a)$.
+We consider this set of algorithms because while they carry the convergence advantages of general policy gradient methods, they also approximate the Q-values with better sample efficiency.
+In our context, we again use neural networks with one hidden layer each for both the actor and critic.
+\begin{algorithm}
+\caption{Q Actor Critic}\label{euclid}
+\begin{algorithmic}[1]
+\Procedure{Actor\_Critic}{}
+\State $\text{Initialize policy and Q-network parameters }\theta, \omega$
+\For {$i = 1, 2, \ldots, M$}
+\State $\text{Initialize state and action }s_1, a_1 \sim \pi_{\theta}(a|s_1)$.
+\For {$t = 1, 2, \ldots, T-1$}
+\State Record reward $r_t$ and next state $s_{t+1}$
+\State Sample the next action $a_{t+1} \sim \log \pi_{\theta}(a|s_{t+1})$
+\State $\text{Update } \theta \leftarrow \theta + \alpha_{\theta} Q_{\omega}(s_t,a_t) \nabla_{\theta} \pi_{\theta} (s_t, a_t)$
+\State $\text{Compute temporal difference error: }\delta_t = r_t + \gamma Q_{\omega}(s_{t+1}, a_{t+1}) - Q_{\omega}(s_{t}, a_{t})$
+\State $\text{Update } \omega \leftarrow \omega + \alpha_{\omega} \delta_t \nabla_{\omega} Q_{\omega} (s_t, a_t)$
+\EndFor
+\EndFor
+\State \Return {$\theta, \omega$}
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+In the current version we have implemented, the
+
+[************** Q-network doesn't converge, maybe because we have an unstable target, requiring a target network]
 

### 2019-12-06 main.py

@@ -18,8 +18,8 @@ def trivial_test():
 
 def reinforce_test():
     m = model.Model(10000)
-    m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.LINEAR)
-    m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.LINEAR)
+    m.create_firms(NUM_FIRMS, rltype=RLType.Q_ACTOR_CRITIC, demand_curve_shape=DemandCurveShape.RECIPROCAL)
+    m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.RECIPROCAL)
     print("Person Skills: %s" % [round(p.skill, 2) for p in m.people])
     num_iters = 100
     avg_GDP = []

### 2019-12-06 analytics.py
new analytics for wealth

@@ -75,19 +75,6 @@ def compute_stats(m, firm_action_hist, person_action_hist,
             GDP_over_time += firm_recv + firm_paid
         stats['GDP_over_time'] = list(GDP_over_time)
 
-        # Gini coefficient over time.
-        p_tot = 0
-        n = 0
-        gini_over_time = np.zeros(len(person_goods_recv[0]))
-        for p1 in people_goods_gained_over_time:
-            p_tot += p1
-            n += 1
-            for p2 in people_goods_gained_over_time:
-                gini_over_time += np.abs(p1 - p2)
-        with np.errstate(divide='ignore', invalid='ignore'):
-            gini_over_time /= 2 * n * p_tot
-        stats['gini_over_time'] = list(gini_over_time)
-
         # Market price of goods.
         goods_tot = 0
         for p in person_goods_recv:
@@ -95,26 +82,61 @@ def compute_stats(m, firm_action_hist, person_action_hist,
         money_tot = 0
         for f in firm_money_recv:
             money_tot += np.array(f)
-        stats['market_price_goods'] = list(money_tot / goods_tot)
+        with np.errstate(divide='ignore', invalid='ignore'):
+            stats['market_price_goods'] = list(money_tot / goods_tot)
 
-        # Market price of labor
+        # Market price of labor.
         hours_tot = 0
         for p in m.people:
             hours_tot += np.array(p.hours_worked)
         money_tot = 0
         for f in firm_money_paid:
             money_tot += np.array(f)
-        stats['market_price_labor'] = list(money_tot / hours_tot)
+        with np.errstate(divide='ignore', invalid='ignore'):
+            stats['market_price_labor'] = list(money_tot / hours_tot)
+
+        # Unemployment.
+        employment = 0
+        for p in m.people:
+            employed_bool = np.array(p.hours_worked) > 0.01
+            employed = [1 if step else 0 for step in list(employed_bool)]
+            employment += np.array(employed)
+        unemployment = 1 - employment / NUM_PEOPLE
+        stats['unemployment'] = list(unemployment)
+
+        # People wealth over time
+        wealth_over_time = []
[…]

### 2019-12-06 analytics.py
thi

@@ -7,9 +7,6 @@ from collections import Counter
 import seaborn as sns
 sns.set(style="darkgrid")
 
-def create_loss_plot(stats, name):
-    pass
-
 def compute_stats(m, firm_action_hist, person_action_hist, 
         firm_money_recv, firm_money_paid, person_goods_recv, person_money_hist,
         firm_money_hist, firm_goods_hist):
@@ -101,10 +98,10 @@ def compute_stats(m, firm_action_hist, person_action_hist,
         stats['market_price_goods'] = list(money_tot / goods_tot)
 
         # Market price of labor
-        hours_tot = 0
+        hours_tot = 0.0
         for p in m.people:
             hours_tot += np.array(p.hours_worked)
-        money_tot = 0
+        money_tot = 0.0
         for f in firm_money_paid:
             money_tot += np.array(f)
         stats['market_price_labor'] = list(money_tot / hours_tot)

### 2019-12-06 main.py

@@ -18,12 +18,39 @@ def trivial_test():
 
 def plot1_for_paper():
     m = model.Model(1e5)
-    m.create_firms(1, rltype=RLType.REINFORCE)
-    m.create_people(5, rltype=RLType.TRIVIAL)
+    m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.RECIPROCAL)
+    m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.RECIPROCAL)
     num_iters = 100
+    firm_loss = []
+    mean_person_loss = []
+    prices = []
+    avg_profit = []
     for i in range(num_iters):
-        stats = m.run_episode(200, verbose=True)
+        stats, firm_losses, person_losses = m.run_episode(200, verbose=True)
+        firm_loss.append(np.mean(firm_losses))
+        mean_person_loss.append(np.mean(person_losses))
         
+        
+        avg_profit.append(np.mean(stats['firm_avg_profit']))
+
+        price_history = stats['firm_price_hists'][0]
+        price_history = sorted([(k, price_history[k]) for k in price_history])
+        prices.append(price_history)
+        print(price_history)
+
+    with open('plot1_for_paper.txt', 'w') as f:
+        f.write(','.join([str(round(x, 2)) for x in firm_loss]) + '\n')
+        f.write(','.join([str(round(x, 2)) for x in mean_person_loss]) + '\n')
+        f.write(','.join([str(round(x, 2)) for x in avg_profit]) + '\n')
+        
+        for price_hist in prices:
+            f.write(','.join([
+                '(' + str(round(x[0], 2)) + ';' + str(round(x[1], 2)) + ')' for x in price_hist
+            ]) + '\n')
+    
+
+
+
 
 def reinforce_test():
     m = model.Model(10000)
@@ -43,7 +70,8 @@ def reinforce_test():
     # plot_wealth_histories(m, stats)
 
 def main():
-    reinforce_test()
+    # reinforce_test()
+    plot1_for_paper()
 
 if __name__ == "__main__":
     main()

### 2019-12-06 model.py

@@ -74,9 +74,6 @@ class Model:
             person_losses.append(person.get_loss())
             person.reset()
 
-        stats['firm_losses'] = firm_losses
-        stats['person_losses'] = person_losses
-
         if verbose:
             # for stat in stats:
             #     print(stat)
@@ -86,7 +83,7 @@ class Model:
             losses = (np.mean(person_losses), np.mean(firm_losses), np.mean(np.concatenate((person_losses, firm_losses))))
             print("Losses (P, F, T) = (%8.5f, %8.5f, %8.5f)" % losses)
 
-        return stats
+        return stats, firm_losses, person_losses
 
     def run(self, num_timesteps=100):
         for _ in range(num_timesteps): self.run_one_step()

### 2019-12-06 person.py
Initial plots for methods section of paper

@@ -10,7 +10,7 @@ class PersonAgent(Agent):
         self.epis_actions = []
         self.goods_recv = []
         self.hours_worked = []
-        self.money_hist = []
+        self.money_hist = [self.init_money]
         self.iteration = 0
 
         # Per-turn state.

### 2019-12-06 firm.py
Changed actor critic to a2c

@@ -109,7 +109,7 @@ class FirmAgent(Agent):
         if self.rltype == RLType.REINFORCE:
             loss_hist = self.policy_net.loss_hist
         elif self.rltype == RLType.Q_ACTOR_CRITIC:
-            loss_hist = list(zip(self.actor_critic.policy_loss_hist, self.actor_critic.q_loss_hist))
+            loss_hist = list(zip(self.actor_critic.policy_loss_hist, self.actor_critic.adv_loss_hist))
 
         if len(loss_hist) == 0: return None
         return loss_hist[-1]

### 2019-12-06 main.py

@@ -70,8 +70,8 @@ def reinforce_test():
     # plot_wealth_histories(m, stats)
 
 def main():
-    # reinforce_test()
-    plot1_for_paper()
+    reinforce_test()
+    #plot1_for_paper()
 
 if __name__ == "__main__":
     main()

### 2019-12-06 model.py

@@ -80,8 +80,9 @@ class Model:
             #     pp(stats[stat])
             #     print("=======")
 
-            losses = (np.mean(person_losses), np.mean(firm_losses), np.mean(np.concatenate((person_losses, firm_losses))))
-            print("Losses (P, F, T) = (%8.5f, %8.5f, %8.5f)" % losses)
+            # losses = (np.mean(person_losses), np.mean(firm_losses), np.mean(np.concatenate((person_losses, firm_losses))))
+            # print("Losses (P, F, T) = (%8.5f, %8.5f, %8.5f)" % losses)
+            print('{}, {}'.format(firm_losses[0][0], firm_losses[0][1]))
 
         return stats, firm_losses, person_losses
 

### 2019-12-06 q_actor_critic.py

@@ -35,13 +35,17 @@ class PolicyNetwork(nn.Module):
         return F.log_softmax(model(x), dim=0)
 
 
-class QNetwork(nn.Module):
-    def __init__(self, state_dim, action_dim, learning_rate=0.01):
-        super(QNetwork, self).__init__()
+class AdvantageNetwork(nn.Module):
+    def __init__(self, state_dim, action_dim, learning_rate=0.005):
+        super(AdvantageNetwork, self).__init__()
         # Should be (2 * number of firms) + (number of people)
         self.state_dim = state_dim 
         # Should be (choices of hours/goods produced) * (choices of prices) * (demand curve parameters)
+        
+
         self.action_dim = action_dim
+
+
 
         self.learning_rate = learning_rate
 
@@ -61,10 +65,10 @@ class QNetwork(nn.Module):
 class ActorCritic(object):
     def __init__(self, state_dim, action_dim):
         self.policy_net = PolicyNetwork(state_dim, action_dim)
-        self.q_net = QNetwork(state_dim, action_dim)
+        self.adv_net = AdvantageNetwork(state_dim, 1)
 
         self.policy_optimizer = optim.Adam(self.policy_net.parameters(), lr=self.policy_net.learning_rate)
-        self.q_optimizer = optim.Adam(self.q_net.parameters(), lr=self.q_net.learning_rate)
+        self.adv_optimizer = optim.Adam(self.adv_net.parameters(), lr=self.adv_net.learning_rate)
 
         self.gamma = DISCOUNT
 
@@ -78,7 +82,7 @@ class ActorCritic(object):
         self.next_reward = None
 
         self.policy_loss_hist = []
-        self.q_loss_hist = []
+        self.adv_loss_hist = []
 
 
     def choose_action(self, state):
@@ -94,8 +98,9 @@ class ActorCritic(object):
     def update_policy(self):
         if type(self.curr_state) == type(None):
             return
-        q_value = torch.index_select(self.q_net(self.curr_state), 0, self.curr_action)
-        policy_loss = torch.mul(self.curr_action_log_prob, q_value).mul(-1)
+        #adv_value = torch.index_select(self.adv_net(self.curr_state), 0, self.curr_action)
+        adv_value = torch.sum(self.adv_net(self.curr_state))
+        policy_loss = torch.mul(self.curr_action_log_prob, adv_value).mul(-1)
         self.policy_optimizer.zero_grad()
         policy_loss.backward()
[…]

### 2019-12-06 analytics.py
writing data to text files

@@ -128,7 +128,59 @@ def compute_stats(m, firm_action_hist, person_action_hist,
             gini_over_time /= 2 * n * p_tot
         stats['gini_over_time'] = list(gini_over_time)
 
+        # Median wealth over time.
+        w = wealth_over_time
+        median_p = [np.median([p[i] for p in w]) for i in range(len(w[0]))]
+        wealth_rel_median = []
+        with np.errstate(divide='ignore', invalid='ignore'):
+            for p in w:
+                wealth_rel_median.append((p - median_p) / median_p)
+        stats['wealth_rel_median'] = wealth_rel_median
+
         return stats
+
+def write_plots_to_file(m, stats, iteration_num, path):
+    cols = [
+    ('people_money_over_time', 'mp*'),
+    ('firm_money_hists', 'mf*'),
+    ('people_goods_over_time', 'gp*'),
+    ('firm_goods_hists', 'gf*'),
+    ('GDP_over_time', 'gdp'),
+    ('unemployment', 'u'),
+    ('market_price_goods', 'pg'),
+    ('market_price_labor', 'pl'),
+    ('wealth_rel_median', 'wp*'),
+    ('gini_over_time', 'gini')
+    ]
+
+    filename = path + str(iteration_num) + '.txt'
+    with open(filename, 'w') as f:
+        # header
+        for col in cols:
+            _, cid = col
+            if cid[-1] == '*':
+                if cid[-2] == 'p':
+                    N = NUM_PEOPLE
+                else:
+                    N = NUM_FIRMS
+                for i in range(N):
+                    f.write(cid[:-1] + str(i) + ' ')
+            else:
+                f.write(cid + ' ')
+        f.write('\n')
+
+        # data
+        n_steps = len(stats['GDP_over_time'])
+        for step in range(n_steps):
+            for col in cols:
+                key, cid = col
+                stat = stats[key]
+                if cid[-1] == '*':
+                    for agent in stat:
+                        f.write(str(round(agent[step], 2)) + ' ')
+                else:
+                    f.write(str(round(stat[step], 2)) + ' ')
+            f.write('\n')
 
 def save_plots_from_iteration(stats, iteration_num, name):
     _, axs = plt.subplots(4, 3, figsize=(15, 12))
@@ -138,7 +190,6 @@ def save_plots_from_iteration(stats, iteration_num, name):
[…]

### 2019-12-06 paper/paper.tex

@@
 \usepackage{subcaption}
+\usepackage{pgfplotstable}
+\usepackage{pgfplots}
 
@@
 \renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
-\renewcommand\thesubsection{\roman{subsection}} % roman numerals for subsections
+\renewcommand\thesubsection{\Alph{subsection}} % roman numerals for subsections
 \titleformat{\section}[block]{\large\scshape\centering}{\thesection.}{1em}{} % Change the look of the section titles
@@
 \fancyfoot{} % Blank out the default footer
-\fancyhead[C]{Modeling an Economy with Policy Gradients $\bullet$ Autumn 2019 $\bullet$ CS 238} % Custom header text
+\fancyhead[C]{Modeling Macroeconomic Phenomena with Multi-Agent Reinforcement Learning $\bullet$ Autumn 2019 $\bullet$ CS 238} % Custom header text
 \fancyfoot[RO,LE]{\thepage} % Custom footer text
@@
 \posttitle{\end{center}} % Article title closing formatting
-\title{Modeling an Economy with Policy Gradients} % Article title
+\title{Modeling Macroeconomic Phenomena \\with Multi-Agent Reinforcement Learning} % Article title
 \author{%
@@
 \begin{abstract}
-
-TODO
-
+While reinforcement learning has been widely used to model gameplay, its application to economics remains limited.
+In this work, we demonstrate how multi-agent reinforcement learning can be combined with a simple economic model to simulate and analyze macroeconomic phenomena under various conditions.
+In particular, we show how policy gradient learning algorithms yield qualitatively correct results in simple simulations, and we demonstrate nontrivial behavior in more complex scenarios involving over 30 separate agents.
 \end{abstract}
@@
 \end{figure}
-
+\medskip
 While the circular flow model does not include many important economic factors, it has the benefit of being simple enough to serve as a starting point for high-level analysis.
@@
 In the following work, we explore how various economic phenomena can be simulated and analyzed by applying reinforcement learning to the circular flow model.
-While we are limited by the simplicity of the model, there are many economic metrics that can be directly investigated in such a model.
+While we are limited by the simplicity of the model, there are many economic metrics that manifest in such a model.
 For example, the gross domestic product (GDP) is widely touted as a comprehensive measure of a country's economic health.
@@
 While the circular flow model cannot account for net exports or government spending, the total consumer spending and private domestic investments can be approximated by the total quantity of money exchanged in the goods and labor markets.
-Thus, such a model can lead to investigation of how GDP changes over time and as a function of the number of agents of each type.
 
@@
 
-Another important economic focus is on income and wealth inequality, both of which are manifested in the circular flow model.
-A common approach to measuring inequality is to use the Gini coefficient.
-The Gini coefficient of a distribution $p$ with mean $\mu$ is given by
+Another important area of economics studies income and wealth inequality.
+A common approach to measuring these is to use the Gini coefficient.
+The Gini coefficient is a measure of the inequality of a distribution $p$ with mean $\mu$.
+Explicitly, the Gini coefficient is given by
 \[
@@
 \] 
-In a discrete scenario, this just becomes the average absolute difference of all pairs of items (e.g., income or absolute wealth) divided by twice the mean.
-Given its relevance to modern society, understanding how income inequality develops and progresses over time and as a function of initial wealth and skill distribution is quite important.
+In a discrete scenario, this becomes the average absolute difference of all pairs of items (e.g., income or absolute wealth) divided by twice the mean.
 
@@
 We aim to explore the applicability of multi-agent reinforcement learning as a means for analyzing macroeconomic phenomena.
-To this end, we implement the circular flow model and perform basic analyses of both GDP and income inequality.
+To this end, we implement the circular flow model and perform basic analyses of several economic metrics, including GDP and the Gini coefficient.
 As this is a multi-agent problem, agents choose actions and learn simultaneously, making the simulation dynamics quite complex.
-Thus, we will also discuss and analyze the effectiveness of two different learning algorithms.
+We also discuss and analyze the effectiveness of two different learning algorithms.
 
@@
 \cite{RePEc:col:000092:003907} which focuses on modeling an economy ``based on conventions.'' In this work, the authors apply the $\text{SARSA}(\lambda)$ algorithm in a government agent to search for a good expenditure policy.
-In this case, the authors avoid applying complicated multi-agent reinforcement learning by modeling each firm with a simple algorithm in which firms can copy neighboring firms or (with some low probability) switch to a random policy.
+The authors avoid applying complicated multi-agent reinforcement learning by modeling each firm with a simple algorithm in which firms can copy neighboring firms or (with some low probability) switch to a random policy.
 The government attempts to increase national income and decrease debt based on some limited knowledge of the existing firms and historical records of income, debt, etc.
@@
 One important issue that the authors of \cite{RePEc:col:000092:003907} grappled with was the vast (and continuous) state space.
-In this case, they applied a multilayer-perceptron (MLP) with a single hidden layer to approximate the $Q$-function for $\text{SARSA}(\lambda)$.
+They applied a multilayer-perceptron (MLP) with a single hidden layer to approximate the $Q$-function for $\text{SARSA}(\lambda)$.
 Given the vast array of actions and states within any economy, approximation with an MLP model or similar is common in such scenarios.
-The authors found limited success with their model, with the economy collapsing less frequently over time due to the government's policies.
+The authors found some success with their model, with the economy collapsing less frequently over time due to the government's policies.
 
@@
 In particular, Tesauro \& Kephart consider two selling agents competing in an economic model where each agent applied Q-learning.
-In this case, the state and action space was small enough that the Q-learning could proceed directly.
+The state and action space was small enough that the Q-learning could proceed directly.
 Furthermore, the consumers in the model acted via a simple, greedy rule as opposed to acting as an independent, market-shaping force.
@@
 To address this, the author suggested applying an MLP or other generalizable model as an extension.
-In our model, we attempt to combine and extend many of the successful techniques presented in past studies to create an accurate representation of the circular flow model.
+The suggestion to combine policy gradient algorithms and a generalizable action-choosing model significanlty influenced our design.
 
 \section{Methods}
+
+The model consists of an implementation of the circular flow architecture and a reinforcement learning engine. We describe these two components separately.
+
+\subsection{Circular Flow Architecture}
 
@@
 \end{itemize}
-Each firm receives a reward at each step equal to the profit they generate.
-Thus, firms aim to maximize their future profit.
+Human agents are additionally each seeded with a constant skill level which governs how many goods a human agent can produce in an hour of work.
+
+\medskip
+
+On each iteration, the internal state of each agent can change, and this change determines the reward for that agent.
 Human agents receive a reward equal to the number of goods they consume.
-Furthermore, human agents are each seeded with a constant skill level which governs how many goods a human agent can produce in an hour of work.
+Firm agents receive a reward equal to the profit increase in logarithmic space.
+More precisely, the reward for a firm is
+\[
+  r_\text{firm} = \log (m + p + \epsilon) - \log(m + \epsilon) = \log\left(1 + \frac{p}{m + \epsilon}\right).  
+\]
+Here $m$ is the money of the firm before the update, $p$ is the net profit of the firm for this iteration, and $\epsilon$ is a small constant used to prevent infinite negative reward in the case of $m = 0$ or $m = p = 0$.
+This definition is derived from the idea that the utility of money is logarithmic.
 
@@
   \item \textbf{Demand Curve.} The demand curve is a vector.
-In the labor market, the $i$th entry of the demand curve corresponds to the price the firm will pay to produce its $i$th good.
+In the labor market, the $i$th entry of the demand curve corresponds to the price a firm will pay to produce its $i$th good.
 In the goods market, the $i$th entry of the demand curve corresponds to the price a human will pay for its $i$th good.
@@
 
-In practice, this action space is vast.
-In order to make the problem tractable, the offered price and offered units are discretized into one of 20 or so values.
-The demand curve is parameterized by TODO.
+The space of all actions is vast.
+In order to make the problem tractable, the offered price and offered units are discretized into $P$ and $U$ values, respectively.
+In our simulations, we often set $P \approx 15$ and $U \approx 25$.
+Similarly, the demand curve was parameterized by families of functions.
+We experimented with two different families of curves: inverse and linear.
+The family of inverse curves $y = a/x + b$ is specified by the two parameters $a, b$ (in our simulations we used approximately 36 total possible configurations).
+The family of linear curves was specified by solely the $y$-intercept, as the $x$-intercept was a hyperparameter fixed before training (the $x$-intercept corresponds to the maximum number of goods that can be purchased or produced by a given human or firm in a single iteration).
+In practice, the linear family trained faster, but the inverse family led to better results.
+We use the family of inverse demand curves for all plots presented in this paper.
 
-[multiagent -> individual RL agents]
-[describe state, $s_s$]
-[describe actions, $s_a$, as discretized to more easily track the]
-[describe rewards]
-[parameterizing the demand curve]
-[full transparency of staters]
+\medskip
+
+The state used to produce the action is the money and goods of all firms and the money of all people (concatenated into a single vector).
+For a system with $N$ firms and $M$ people, the resulting state space has dimension $s_s = 2N + M$.
+Thus, the reinforcement learning engine (described in the next section) is tasked with learning a mapping from a space of size $s_s$ to a space of size $s_a \approx P \cdot U \cdot F$, where $F$ is the number of possible values for the parameters of the demand curves.
+While this fully transparent state is not wholly realistic, it significantly simplifies the model and can be justified economically by the increasing availability of information.
+
+\medskip
+
+In addition to this setup, several additional features were added to enhance the simulations.
+All human agents' stored money is increased every iteration by an interest rate (generally $<1\%$).
+This has the benefit of preventing trivial policies in which humans provide all money to the firms, at which point the firms can no longer directly increase reward.
+Similarly, every iteration firms' stored money is decreased by fixed operational costs as well as costs that scale with the size of the firm (i.e., $m \mapsto r(m - f)$, where $f$ is a fixed cost and $r$ is a multiplicative decay just below $1$).
+Together, these additions allow for interesting model dynamics for the entirety of the experiment.
 
 \subsection{RL Algorithms******}
-Following the literature we reviewed, we worked within the realm of policy gradient methods; we also choose this family of algorithms because they have stronger convergence guarantees than value-based RL methods (albeit potentially to local optimum that are not global optimum).
-\newline\newline
-As framed, our problem contains large — in fact, high dimensional — state spaces.
-Specifically, the dimension of a given state is $s_s = 2 \cdot (\text{number of firms}) + (\text{number of people})$.
-Furthermore, while the values in each dimension were integers, they could take on a wide range of values.
-Therefore, in every policy, we use neural network architectures to convert a given state to a vector, containing log probabilities of taking each action.
+Both the learning algorithms chosen for the simulation were policy gradient methods.
+This was chosen due to previous literature, and because such methods have stronger convergence guarantees than value-based reinforcement learning methods (though potentially to non-global optima).
+
+\medskip
+
+As discussed previously, the simulation contains large, high-dimensional state spaces.
+The dimension of a given state vector is $s_s = 2 N + M$.
+Furthermore, every entry can take on a wide range of integral values.
+Therefore, we use neural network architectures to convert a given state to a vector containing log probabilities for each action.
 In empirical testing, if $s_a$ is the number of possible actions, we've seen that the following architecture is a simple framework that still yields convergence in our context:
@@
 \end{itemize}
-During training, we generally use a learning rate of $\alpha = 0.01$.
+During training, the learning rate was set to $\alpha = 0.01$.
 
@@
 Perhaps the simplest policy gradient method is the REINFORCE algorithm.
-This algorithm uses a Monte Carlo-type approach to gathering training data.
-We manually specify the length of the episode, generally using $T = 100$.
+This algorithm uses a Monte Carlo approach to collecting training data.
+The length of the episode $T$ is manually specified, and we generally used $T \sim 100$.
 \begin{algorithm}
@@
 \end{algorithm}
-This algorithm runs reasonably quickly per episode. However, we also need to run more episodes because:
+The runtime of this algorithm per episode is relatively short. However, often a significant number of episodes are required for the following reasons:
 \begin{itemize}
@@
   \item In many contexts, vanilla REINFORCE has been known to suffer from high variance in the losses from episode to episode.
-  \item In our multi-agent problem, the environment dynamics constantly change because every agent responds to each timestep in the markets.
+  \item In our multi-agent problem, the environment dynamics constantly change as every agent responds to each timestep in the markets.
 \end{itemize}
@@
 An alternative family of policy gradient methods is the \textit{actor-critic} family.
-In these algorithms, the actor component computes the log probabilities $\log \pi_{\theta}(s, a)$ while the critic component approximates the functions $Q(s, a)$.
-We consider this set of algorithms because while they carry the convergence advantages of general policy gradient methods, they also approximate the Q-values with better sample efficiency.
-In our context, we again use neural networks with one hidden layer each for both the actor and critic.
+In these algorithms, the actor component computes the log probabilities $\log \pi_{\theta}(s, a)$ while the critic component approximates the function $Q(s, a)$.
+We considered this set of algorithms because it carries the convergence advantages of general policy gradient methods while also approximating Q-values with higher sample efficiency.
+As with REINFORCE, we again used neural networks with one hidden layer for both the actor and critic.
 \begin{algorithm}
@@
 
+We break the results down into two sections: preliminary results and a more complex simulation.
+The preliminary results are designed to verify that the system is working and producing correct results, while the more complex simulation allows more open-ended exploration of economic concepts.
+
+\subsection{Preliminary Results}
+
+To test the convergence of the learning and the ensure that the results were logical, we created a scenario with one firm agent and five human agents.
+The dynamics discussed below are for a simulation in which all agents utilized the REINFORCE algorithm, all agents used inverse demand curves, the human agent interest rate was set to $1\%$, the firm operational cost was set to $\$10$, and the discount was $\gamma = 0.99$.
+The firm was initialized with $\$10000$, while the human agents were initialized to have uniform random initial money such that the total money given to human agents was $\$5000$.
+The human skill levels (goods per hour) were drawn from $N(1, 0.1)$.
+The simulation was run for 100 iterations.
+
+\medskip
+
+Recall that the loss function for the REINFORCE algorithm is of the following form:
+\[
+  \mathcal{L} = -\sum_{t = 1}^T \left[\log \pi_\theta(s_t, a_t) \left( \sum_{i = t}^n \gamma^{i - t}r_i \right) \right].  
+\]
+
+In particular, as the policy converges, the probability of taking a given action $a_t$ from state $s_t$ should approach one, and so the $\log \pi_\theta$ term will approach zero.
+Therefore, convergence corresponds to the losses approaching zero.
+This occurs in our preliminary simulation, as can be seen in figure \ref{plt:losstozero}.
+
+\begin{figure}[h]
+\begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+  \begin{axis}[
+    xlabel=Episode,
+    ylabel=Loss,
+    legend style={at={(1,0)}, anchor=south east}
+  ]
+  \addplot table [y=fl, x=i]{prelims.dat};
+  \addlegendentry{Firm}
+  \addplot table [y=pl, x=i]{prelims.dat};
+  \addlegendentry{Human (Mean)}
+  \end{axis}
+\end{tikzpicture}
+\caption{The agents' losses converge to zero over time.
+This plot illustrates the loss of firm agents vs.
+human agents in a simulation with one firm agent and five human agents.
+The loss for the human agents is the average of all five human agents' losses.
+This simulation uses the REINFORCE learning algorithm, and indicates that the algorithm is succcessfully converging to a policy within 100 episodes.}
+\label{plt:losstozero}
+\end{figure}
+
+While this indicates that agents are able to converge to policies in this multi-agent environment, it does not necessarily indicate that the learned behavior leads to positive rewards.
+To investigate this, we can look at how the firm's profit changes over time and the actions that the firm is taking.
+As there is a single firm and multiple people, we might expect that the firm has more of an ability to monopolize the market, pushing the price of goods higher.
+As such, we expect the price at which the firm sells goods to rise.
+In figure \ref{plt:firmprofitincrease}, we see that the firm's profit is generally increasing over time from an initial average profit of approximately $\$23$ per iteration to an average profit of up to $\$50$ per iteration.
+The noisiness in the increase in profit may be due to the complex, changing environment (due to the changing actions of the human agents), or the fact that the reward is actually the future-discounted logarithmic profit.
+
+\medskip
+
+Figure \ref{plt:firmactions} illustrates the frequency at which the firm chooses a given price in the first and last episodes.
+In particular, the firm generally learns to select higher prices as it has a monopoly on the goods market and the five human agents are competing for the firm's offerings.
+Note that this precise pattern is not wholly reproducible; on subsequent runs, the firm agent will display similar behavior but the precise actions chosen will vary.
+
+\begin{figure}[h]
+  \begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+    \begin{axis}[
+      xlabel=Episode,
+      ylabel=Average Firm Profit per Iteration
+    ]
+    \addplot table [y=fp, x=i]{prelims.dat};
+    \end{axis}
+  \end{tikzpicture}
+  \caption{Firm profit increases over time.
+This plot shows the average profit the firm gains per iteration as a function of episode.
+The simulation shown uses a single firm agent and five human agents, all using the REINFORCE learning algorithm.
+This indicates that the REINFORCE learning algorithm is yielding higher reward and better policies over time.}
+  \label{plt:firmprofitincrease}
+  \end{figure}
+
+\begin{figure}[h]
+\begin{tikzpicture}
+    \begin{axis}[
+        ybar,
+        legend style={at={(0.5,-0.2)},
+          anchor=north,legend columns=-1},
+        ylabel=Frequency,
+        bar width=5pt,
+        xlabel=Price per Good (\$),
+        symbolic x coords={1, 4, 7, 10, 13, 16, 19, 22, 25, 28},
+        xtick=data
+        ]
+    \addplot coordinates {(1, 19) (4, 22) (7, 20) (10, 27) (13, 15) (16, 27) (19, 16) (22, 17) (25, 14) (28, 23)};
+    \addplot coordinates {(1, 5) (4, 21) (7, 25) (10, 0) (13, 0) (16, 0) (19, 0) (22, 106) (25, 0) (28, 43)};
+    \legend{episode 1, episode 100}
+    \end{axis}
+  \end{tikzpicture}
+  \caption{With a single firm agent and five human agents, the firm learns to increase the price of goods to maximize its own profit.
+The bars shown correspond to the frequency at which the firm chooses a given price to offer for the first episode and the last episode.
+The possible prices are discretized to be in the set $\{1, 4, 7, 10, 13, 16, 19, 22, 25, 28\}$.
+This behavior is consistent with that of a monopoly.}
+  \label{plt:firmactions}
+\end{figure}
+
+\subsection{Investigating Economic Phenomena}
+
 \section{Conclusion}

### 2019-12-06 main.py
plotting for a2c

@@ -18,16 +18,17 @@ def trivial_test():
 
 def plot1_for_paper():
     m = model.Model(1e5)
-    m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.RECIPROCAL)
+    m.create_firms(NUM_FIRMS, rltype=RLType.Q_ACTOR_CRITIC, demand_curve_shape=DemandCurveShape.RECIPROCAL)
     m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.RECIPROCAL)
-    num_iters = 100
+    num_iters = 50
     firm_loss = []
     mean_person_loss = []
     prices = []
     avg_profit = []
     for i in range(num_iters):
-        stats, firm_losses, person_losses = m.run_episode(200, verbose=True)
-        firm_loss.append(np.mean(firm_losses))
+        stats, firm_losses, person_losses = m.run_episode(300, verbose=True)
+        #firm_loss.append(np.mean(firm_losses))
+        firm_loss.append(firm_losses[0])
         mean_person_loss.append(np.mean(person_losses))
         
         
@@ -38,15 +39,28 @@ def plot1_for_paper():
         prices.append(price_history)
         print(price_history)
 
-    with open('plot1_for_paper.txt', 'w') as f:
-        f.write(','.join([str(round(x, 2)) for x in firm_loss]) + '\n')
-        f.write(','.join([str(round(x, 2)) for x in mean_person_loss]) + '\n')
-        f.write(','.join([str(round(x, 2)) for x in avg_profit]) + '\n')
+    fig, axs = plt.subplots(2, 1, figsize=(8, 12))
+
+    fig.suptitle('Actor-Critic Losses with Firm Monopoly', fontsize=20)
+    axs[0].plot([f[0] for f in firm_loss])
+    axs[0].set_xlabel("Episode")
+    axs[0].set_ylabel("Policy Network Loss", fontsize=16)
+    axs[1].plot([f[1] for f in firm_loss])
+    axs[1].set_xlabel("Episode")
+    axs[1].set_ylabel("Advantage Network Loss", fontsize=16)
+    fig.subplots_adjust(top=0.95)
+    plt.savefig('actor_critic_losses.png', bbox_inches='tight', dpi=300)
+    plt.close()
+
+    # with open('plot1_for_paper.txt', 'w') as f:
[…]

### 2019-12-06 model.py

@@ -80,9 +80,13 @@ class Model:
             #     pp(stats[stat])
             #     print("=======")
 
-            # losses = (np.mean(person_losses), np.mean(firm_losses), np.mean(np.concatenate((person_losses, firm_losses))))
-            # print("Losses (P, F, T) = (%8.5f, %8.5f, %8.5f)" % losses)
-            print('{}, {}'.format(firm_losses[0][0], firm_losses[0][1]))
+            #losses = (np.mean(person_losses), np.mean(firm_losses), np.mean(np.concatenate((person_losses, firm_losses))))
+            #print("Losses (P, F, T) = (%8.5f, %8.5f, %8.5f)" % losses)
+
+            losses = (np.mean(person_losses), np.mean([f[0] for f in firm_losses]), np.mean([f[1] for f in firm_losses]))
+            print("Losses (P, FP, FA) = (%8.5f, %8.5f, %8.5f)" % losses)
+
+            # print('{}, {}'.format(firm_losses[0][0], firm_losses[0][1]))
 
         return stats, firm_losses, person_losses
 

### 2019-12-06 q_actor_critic.py

@@ -36,7 +36,7 @@ class PolicyNetwork(nn.Module):
 
 
 class AdvantageNetwork(nn.Module):
-    def __init__(self, state_dim, action_dim, learning_rate=0.005):
+    def __init__(self, state_dim, action_dim, learning_rate=0.01):
         super(AdvantageNetwork, self).__init__()
         # Should be (2 * number of firms) + (number of people)
         self.state_dim = state_dim 
@@ -141,13 +141,16 @@ class ActorCritic(object):
         self.next_action_log_prob = None
         self.next_reward = None
 
-        if len(self.policy_loss_hist) > 0:
-            print('Current losses: {} {}'.format(self.policy_loss_hist[-1], self.adv_loss_hist[-1]))
+        # if len(self.policy_loss_hist) > 0:
+        #     print('Current losses: {} {}'.format(self.policy_loss_hist[-1], self.adv_loss_hist[-1]))
 
     def record_reward(self, reward):
         self.next_reward = reward
 
     def reset_memory(self):
+
+        print('Last Advantage: {}'.format(torch.sum(self.adv_net(self.curr_state))))
+
         self.curr_state = None
         self.curr_action = None
         self.curr_action_log_prob = None

### 2019-12-06 paper/paper.tex

@@
 
-\subsubsection{Q Actor-Critic}
+\subsubsection{Actor-Critic}
 An alternative family of policy gradient methods is the \textit{actor-critic} family.
-In these algorithms, the actor component computes the log probabilities $\log \pi_{\theta}(s, a)$ while the critic component approximates the function $Q(s, a)$.
-We considered this set of algorithms because it carries the convergence advantages of general policy gradient methods while also approximating Q-values with higher sample efficiency.
+In these algorithms, the actor component computes the log probabilities $\log \pi_{\theta}(s, a)$ while the critic component approximates the performance of the action.
+We considered this set of algorithms because in theory, it carries the convergence advantages of general policy gradient methods while also approximating Q-values with higher sample efficiency.
+In this scenario, we tested Q Actor-Critic, in which the critic computes the Q-values $Q(s,a)$, as well as Advantage Actor-Critic (A2C), in which the critic computes the advantage $Q(s, a) - V(s)$; the idea of using the advantage vs the $Q$-value is to compare the quality of the action with a stable baseline, reducing variance.
 As with REINFORCE, we again used neural networks with one hidden layer for both the actor and critic.
+Note also that we get to update the parameters after every action via TD learning.
 \begin{algorithm}
@@
 \end{algorithm}
-In the current version we have implemented, the
-
-[************** Q-network doesn't converge, maybe because we have an unstable target, requiring a target network]
 
@@
 
+\begin{figure}[h]
+\begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+  \begin{axis}[
+    xlabel=Episode,
+    ylabel=Policy Loss,
+    legend style={at={(1,0)}, anchor=south east}
+  ]
+  \addplot table [y=pl, x=i]{a2c.dat};
+  \addlegendentry{Firm}
+  \end{axis}
+\end{tikzpicture}
+\caption{Policy Network Loss in Actor Critic Method}
+\label{plt:a2cpolicyloss}
+\end{figure}
+
+\begin{figure}[h]
+\begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+  \begin{axis}[
+    xlabel=Episode,
+    ylabel=Critic Loss,
+    legend style={at={(1,0)}, anchor=south east}
+  ]
+  \addplot table [y=al, x=i]{a2c.dat};
+  \addlegendentry{Firm}
+  \end{axis}
+\end{tikzpicture}
+\caption{Critic Network Loss in Actor Critic Method}
+\label{plt:a2cadvloss}
+\end{figure}
+
 \subsection{Investigating Economic Phenomena}

### 2019-12-06 analytics.py
pushing in case of battery failure

@@ -156,6 +156,7 @@ def write_plots_to_file(m, stats, iteration_num, path):
     filename = path + str(iteration_num) + '.txt'
     with open(filename, 'w') as f:
         # header
+        f.write('step ')
         for col in cols:
             _, cid = col
             if cid[-1] == '*':
@@ -170,16 +171,18 @@ def write_plots_to_file(m, stats, iteration_num, path):
         f.write('\n')
 
         # data
+        acc = 4
         n_steps = len(stats['GDP_over_time'])
         for step in range(n_steps):
+            f.write(str(step) + ' ')
             for col in cols:
                 key, cid = col
                 stat = stats[key]
                 if cid[-1] == '*':
                     for agent in stat:
-                        f.write(str(round(agent[step], 2)) + ' ')
+                        f.write(str(round(agent[step], acc)) + ' ')
                 else:
-                    f.write(str(round(stat[step], 2)) + ' ')
+                    f.write(str(round(stat[step], acc)) + ' ')
             f.write('\n')
 
 def save_plots_from_iteration(stats, iteration_num, name):

### 2019-12-06 main.py

@@ -48,30 +48,20 @@ def plot1_for_paper():
                 '(' + str(round(x[0], 2)) + ';' + str(round(x[1], 2)) + ')' for x in price_hist
             ]) + '\n')
     
-
-
-
-
 def reinforce_test():
     m = model.Model(10000)
-    m.create_firms(NUM_FIRMS, rltype=RLType.Q_ACTOR_CRITIC, demand_curve_shape=DemandCurveShape.RECIPROCAL)
-    m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.RECIPROCAL)
+    m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.LINEAR)
+    m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.LINEAR)
     print("Person Skills: %s" % [round(p.skill, 2) for p in m.people])
     num_iters = 100
-    avg_GDP = []
     for i in range(num_iters):
-        stats = m.run_episode(300, verbose=True)
-        save_plots_from_iteration(stats, i, 'plots/test9/test')
-        # avg_GDP.append(np.mean(stats['GDP_over_time']))
+        stats, _, _ = m.run_episode(100, verbose=True)
+        save_plots_from_iteration(stats, i, 'plots/test_full_4/test')
+        write_plots_to_file(m, stats, i, 'plots/test_full_4/test_')
 
-    stats = m.run_episode(300)
-    save_plots_from_iteration(stats, i, 'plots/test9/test-final')
-    # print_stats(m, stats)
-    # plot_wealth_histories(m, stats)
 
 def main():
-    # reinforce_test()
-    plot1_for_paper()
+    reinforce_test()
 
 if __name__ == "__main__":
     main()

### 2019-12-06 analytics.py
Commit to pull rory changes

@@ -138,7 +138,6 @@ def save_plots_from_iteration(stats, iteration_num, name):
     plot_gdp_smoothed(axs[2, 2], stats)
     plot_median_human_wealth(axs[2, 1], stats)
     plot_total_money(axs[0, 2], stats)
-    plot_gdp_smoothed(axs[1, 0], stats)
     plot_human_goods_hist(axs[1, 1], stats)
     plot_market_price_goods(axs[1, 2], stats)
     plot_gini_coef(axs[2, 0], stats)

### 2019-12-06 constants.py

@@ -4,8 +4,8 @@ import itertools
 
 # These are constants because I use them to initialize 
 # the policy network's input dimension
-NUM_FIRMS = 1
-NUM_PEOPLE = 5
+NUM_FIRMS = 5
+NUM_PEOPLE = 25
 
 NUM_GOODS_MAX_BUY = 10
 WORK_HOURS_PER_PERSON = 10

### 2019-12-06 main.py

@@ -54,24 +54,22 @@ def plot1_for_paper():
 
 def reinforce_test():
     m = model.Model(10000)
-    m.create_firms(NUM_FIRMS, rltype=RLType.Q_ACTOR_CRITIC, demand_curve_shape=DemandCurveShape.RECIPROCAL)
-    m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.RECIPROCAL)
+    m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.LINEAR)
+    m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.LINEAR)
     print("Person Skills: %s" % [round(p.skill, 2) for p in m.people])
     num_iters = 100
-    avg_GDP = []
     for i in range(num_iters):
-        stats = m.run_episode(300, verbose=True)
-        save_plots_from_iteration(stats, i, 'plots/test9/test')
-        # avg_GDP.append(np.mean(stats['GDP_over_time']))
+        stats, _, _ = m.run_episode(200, verbose=True)
+        save_plots_from_iteration(stats, i, 'plots/test9-lin2/test')
 
-    stats = m.run_episode(300)
-    save_plots_from_iteration(stats, i, 'plots/test9/test-final')
+    stats = m.run_episode(200)
+    save_plots_from_iteration(stats, i, 'plots/test9-lin2/test-final')
     # print_stats(m, stats)
     # plot_wealth_histories(m, stats)
 
 def main():
-    # reinforce_test()
-    plot1_for_paper()
+    reinforce_test()
+    # plot1_for_paper()
 
 if __name__ == "__main__":
     main()

### 2019-12-06 paper/paper.tex
Added information + captions about actor critic

@@
 
-\subsection{RL Algorithms******}
+\subsection{RL Algorithms}
 Both the learning algorithms chosen for the simulation were policy gradient methods.
@@
 
+We also attempted to run both Q Actor-Critic and A2C algorithms for the firm in the same monopoly situation as before (one firm with five people).
+However, while the policy networks would consistently converge (see Figure \ref{plt:acpolicyloss}), the critic networks would rarely converge, about 10\% of the time (see Figure \ref{plt:acadvloss}).
+For this reason, we chose to stick with REINFORCE as the reinforcement learning model for future simulations.
+
 \begin{figure}[h]
@@
 \end{tikzpicture}
-\caption{Policy Network Loss in Actor Critic Method}
-\label{plt:a2cpolicyloss}
+\caption{Policy Network Loss in Q Actor Critic Method: with one actor-critic firm agent and five human agents, the firm's policy network loss converges rather quickly to 0 (note that the y-axis increments are rather small).
+Like in REINFORCE, when the policy becomes close to deterministic, the loss approaches zero.}
+\label{plt:acpolicyloss}
 \end{figure}
@@
 \end{tikzpicture}
-\caption{Critic Network Loss in Actor Critic Method}
-\label{plt:a2cadvloss}
+\caption{Critic Network Loss in Q Actor Critic Method: with one actor-critic firm agent and five human agents, the firm's critic loss does not converge, even with changing the learning rate.
+This was also true when we used A2C, which is supposed to address variance issues with a baseline.
+Note that the scale on the y-axis is quite large.}
+\label{plt:acadvloss}
 \end{figure}

### 2019-12-06 paper/paper.tex
Added citations for RL

@@
 \begin{algorithm}
-\caption{REINFORCE Policy Gradient}\label{euclid}
+\caption{REINFORCE Policy Gradient \cite{reinforce-lectures}}\label{euclid}
 \begin{algorithmic}[1]
@@
   \item Agents only update the parameters once per episode.
-  \item In many contexts, vanilla REINFORCE has been known to suffer from high variance in the losses from episode to episode.
+  \item In many contexts, vanilla REINFORCE has been known to suffer from high variance in the losses from episode to episode.\cite{reinforce-lectures}
   \item In our multi-agent problem, the environment dynamics constantly change as every agent responds to each timestep in the markets.
@@
 In these algorithms, the actor component computes the log probabilities $\log \pi_{\theta}(s, a)$ while the critic component approximates the performance of the action.
-We considered this set of algorithms because in theory, it carries the convergence advantages of general policy gradient methods while also approximating Q-values with higher sample efficiency.
+We considered this set of algorithms because in theory, it carries the convergence advantages of general policy gradient methods while reducing the gradient variance \cite{actor-critic}.
 In this scenario, we tested Q Actor-Critic, in which the critic computes the Q-values $Q(s,a)$, as well as Advantage Actor-Critic (A2C), in which the critic computes the advantage $Q(s, a) - V(s)$; the idea of using the advantage vs the $Q$-value is to compare the quality of the action with a stable baseline, reducing variance.

### 2019-12-06 paper/paper.tex
Paper conclusion written

--- 2019-12-06 paper/paper.tex
@@
 
-\usepackage[hmarginratio=1:1,top=25mm,columnsep=20pt,margin=0.75in]{geometry} % Document margins
+\usepackage[hmarginratio=1:1,top=25mm,columnsep=20pt,margin=0.5in, top=0.75in, bottom=0.75in]{geometry} % Document margins
 \usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
@@
 
+We implemented the circular flow model using multi-agent reinforcement learning and demonstrated that the results were qualitatively correct and could be used to investigate more complex macroeconomic phenomena.
+In particular, we found reasonable results in a simple monopolization model with a single firm agent and five human agents.
+This model was tested with two different policy gradient algorithms: REINFORCE and Q-actor critic.
+The latter was faster, but suffered from stability issues as compared to the former.
+Additionally, we simulated a more complex situation involving 5 firm agents and 25 human agents.
+The resulting dynamics had realistic features, including policies that corresponded to retirement behavior, slowly increasing wealth inequality, etc.
+
+\medskip
+
+While this simulation is a successful proof-of-concept for using multi-agent reinforcement learning for  macroeconomic simulations, there are many extensions that can make the results more realistic and interpretable.
+We outline several of these extensions here:
+\begin{itemize}
+  \item \textbf{Government.} In practice, the GDP tended to decrease in our simulations.
+One way to combat this would be to add a government agent that can control the interest rates and taxes for firms and humans.
+The reward for the government agent would be related to increasing GDP, decreasing unemployment, and keeping the Gini coefficient in a desired range.
+  \item \textbf{Multiple Goods.} We only simulated one specific market on one type of good.
+By expanding to $n$ goods, each person could specialize in the good they were the best at (the skill of a person would become a $n$-vector of skills).
+Additionally, firms could either produce multiple goods or a single good.
+  \item \textbf{Birth, Death, Bankruptcy.} In practice, people can die and wealth can be passed on within a family.
+Similarly, firms can go bankrupt and new firms can emerge.
+Adding a system by which firms and people can enter and leave markets would allow for more generalizable dynamics.
+  \item \textbf{Unions and Trusts.} Allowing cooperation amongst firms and humans would make the simulation even more realistic.
+This can be approximated by allowing one neural network to choose the actions for multiple people or multiple firms at once.
+  \item \textbf{Equity and the Stock Market.} One feature missing from the circular flow model is the existence of financial markets.
+One interesting addition would be allowing human agents to bet on the stock market or to own shares of a firm.
+However, this may require increasing the number of agents significantly in order to prevent a single person's actions from influencing the remainder of the market.
+\end{itemize}
+
+As computational power increases and the field of reinforcement learning progresses, the ability to create accurate economic simulations will continue to improve.
+In particular, economics is a field in which controlled experiments are extremely difficult to setup.
+We hope that the techniques presented in this paper may allow future economists to investigate niche phenomena and situations more easily.
+
 \section{Acknowledgments}
@@
 
-The authors would like to thank the CS 238 staff for a really great time throughout this course \texttt{;)}.
+The authors would like to thank the CS 238 staff for a great time throughout this course.
 

### 2019-12-06 main.py

@@ -77,36 +77,20 @@ def reinforce_test():
     m.create_firms(NUM_FIRMS, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.LINEAR)
     m.create_people(NUM_PEOPLE, rltype=RLType.REINFORCE, demand_curve_shape=DemandCurveShape.LINEAR)
     print("Person Skills: %s" % [round(p.skill, 2) for p in m.people])
-<<<<<<< HEAD
     num_iters = 100
     for i in range(num_iters):
-        stats, _, _ = m.run_episode(200, verbose=True)
-        save_plots_from_iteration(stats, i, 'plots/test9-lin2/test')
-
-    stats = m.run_episode(200)
-    save_plots_from_iteration(stats, i, 'plots/test9-lin2/test-final')
-=======
-    num_iters = 10
-    avg_GDP = []
-    for i in range(num_iters):
         stats, _, _ = m.run_episode(300, verbose=True)
-        save_plots_from_iteration(stats, i, 'plots/test_a2c/test')
-        # avg_GDP.append(np.mean(stats['GDP_over_time']))
+        save_plots_from_iteration(stats, i, 'plots/test10-lin2/test')
+        write_plots_to_file(m, stats, i, 'plots/test10-lin2/data')
 
-    stats, _, _ = m.run_episode(300)
-    save_plots_from_iteration(stats, i, 'plots/test_a2c/test-final')
->>>>>>> 88f69a9071557eed7cd48b5bfd00850dc87a0fd7
+    stats = m.run_episode(300)
+    save_plots_from_iteration(stats, i, 'plots/test10-lin2/test-final')
     # print_stats(m, stats)
     # plot_wealth_histories(m, stats)
 
 def main():
-<<<<<<< HEAD
     reinforce_test()
     # plot1_for_paper()
-=======
-    #reinforce_test()
-    plot1_for_paper()
->>>>>>> 88f69a9071557eed7cd48b5bfd00850dc87a0fd7
 
 if __name__ == "__main__":
     main()

### 2019-12-06 model.py

@@ -80,11 +80,11 @@ class Model:
             #     pp(stats[stat])
             #     print("=======")
 
-            #losses = (np.mean(person_losses), np.mean(firm_losses), np.mean(np.concatenate((person_losses, firm_losses))))
-            #print("Losses (P, F, T) = (%8.5f, %8.5f, %8.5f)" % losses)
+            losses = (np.mean(person_losses), np.mean(firm_losses), np.mean(np.concatenate((person_losses, firm_losses))))
+            print("Losses (P, F, T) = (%8.5f, %8.5f, %8.5f)" % losses)
 
-            losses = (np.mean(person_losses), np.mean([f[0] for f in firm_losses]), np.mean([f[1] for f in firm_losses]))
-            print("Losses (P, FP, FA) = (%8.5f, %8.5f, %8.5f)" % losses)
+            # losses = (np.mean(person_losses), np.mean([f[0] for f in firm_losses]), np.mean([f[1] for f in firm_losses]))
+            # print("Losses (P, FP, FA) = (%8.5f, %8.5f, %8.5f)" % losses)
 
             # print('{}, {}'.format(firm_losses[0][0], firm_losses[0][1]))
 

### 2019-12-06 paper/paper.tex
Fixed policy gradient section with explanation

@@
 
+In policy gradient methods, instead of approximating the value functions and using a separate exploration strategy for choosing actions, we directly compute the policy.
+Given a policy model parameterized by $\theta$, we compute probabilities $\pi_{\theta}(a|s)$.
+When training this model, we aim to minimize the loss function \cite{actor-critic}:
+\[
+J(\theta) = -\sum_{s \in S} P_{\pi}(s) \sum_{a\in A} \pi_{\theta}(a|s) Q^{\pi}(s, a)
+\]
+Intuitively, a low loss means that the agent outputs high probabilities for high $Q$-valued actions.
+Policy Gradient Theorem allows us to express the gradient of this loss in a tractable manner \cite{reinforce-lectures} \cite{actor-critic}:
+\[
+\nabla_{\theta} J(\theta) = \mathbb{E}_{\pi}[Q^{\pi}(s, a) \nabla_{\theta} \log \pi_{\theta}(a|s)]
+\]
+This means that over a trajectory, we simply need to compute the gradients:
+\[
+\left(\sum_{i=t}^T \gamma^{i-t} r_t\right) \nabla_{\theta} \log \pi_{\theta}(a_t|s_t)
+\]
+We can then use standard gradient descent methods to optimize $\theta$.
+
+\medskip
+
 As discussed previously, the simulation contains large, high-dimensional state spaces.
@@
 \For {$t = 1, 2, \ldots, T$}
-\State $\text{Compute }v_t = \sum_{i = t}^n \gamma^{i- t} r_i$.
+\State $\text{Compute }v_t = \sum_{i = t}^T \gamma^{i- t} r_i$.
 \State $\text{Update } \theta \leftarrow \theta + \alpha \nabla_{\theta}\log\pi_{\theta} (s_t, a_t) v_t$
@@
 
-Recall that the loss function for the REINFORCE algorithm is of the following form:
+Recall that the gradient the REINFORCE algorithm is of the following form:
 \[
-  \mathcal{L} = -\sum_{t = 1}^T \left[\log \pi_\theta(s_t, a_t) \left( \sum_{i = t}^n \gamma^{i - t}r_i \right) \right].  
+  \nabla_{\theta} J(\theta) = -\sum_{t = 1}^T \left[\nabla_{\theta} \log \pi_\theta(s_t, a_t) \left( \sum_{i = t}^n \gamma^{i - t}r_i \right) \right].  
 \]
 
-In particular, as the policy converges, the probability of taking a given action $a_t$ from state $s_t$ should approach one, and so the $\log \pi_\theta$ term will approach zero.
-Therefore, convergence corresponds to the losses approaching zero.
+In our training, we designate a proxy for the loss to reflect this gradient:
+\[
+  J'(\theta) = -\sum_{t = 1}^T \left[\log \pi_\theta(s_t, a_t) \left( \sum_{i = t}^n \gamma^{i - t}r_i \right) \right].  
+\]
+Note that, as the policy converges, the probability of taking a given action $a_t$ from state $s_t$ should approach one, and so the $\log \pi_\theta$ term will approach zero.
+Therefore, convergence corresponds to the (proxy) losses approaching zero.
 This occurs in our preliminary simulation, as can be seen in figure \ref{plt:losstozero}.
@@
     xlabel=Episode,
-    ylabel=Loss,
+    ylabel=Loss (Proxy),
     legend style={at={(1,0)}, anchor=south east}

### 2019-12-06 paper/paper.tex
Andy's edits to paper

@@
 al.
-found that reinforcement learning alone does not accurately model humans in competitive social environments, but it \textit{does} accurately model this when combined with biased reward representations \cite{vandenBos2137}.
+found that reinforcement learning alone does not accurately model humans (who deviate from rational behavior) in competitive social environments, but it \textit{does} accurately model this when combined with biased reward representations \cite{vandenBos2137}.
 In a separate work, Bereby-Meyer and Roth experiment with reinforcement learning in game theoretic situations, such as the repeated prisoner's dilemma, when complicated by noise \cite{10.1257/aer.96.4.1029}.
@@
 The family of linear curves was specified by solely the $y$-intercept, as the $x$-intercept was a hyperparameter fixed before training (the $x$-intercept corresponds to the maximum number of goods that can be purchased or produced by a given human or firm in a single iteration).
-In practice, the linear family trained faster, but the inverse family led to better results.
+In practice, the linear family trained faster, but the inverse family led to more realistic behavior.
 We use the family of inverse demand curves for all plots presented in this paper.
@@
 In particular, we found reasonable results in a simple monopolization model with a single firm agent and five human agents.
-This model was tested with two different policy gradient algorithms: REINFORCE and Q-actor critic.
-The latter was faster, but suffered from stability issues as compared to the former.
+This model was tested with two different policy gradient algorithms: REINFORCE and actor critic.
+The latter's policy gradients converged faster, but the overall algorithm suffered from stability issues as compared to the former.
 Additionally, we simulated a more complex situation involving 5 firm agents and 25 human agents.

### 2019-12-06 paper/paper.tex
Working on paper

--- 2019-12-06 paper/paper.tex
@@
 In the labor market, firms pay people for their labor to produce more goods.
-This basic model is diagrammed in figure \ref{fig:circularflow}.
+This basic model is diagrammed in Figure \ref{fig:circularflow}.
 The name comes from the fact that currency flows clockwise in a circle between firms and people.
@@
 This has the benefit of preventing trivial policies in which humans provide all money to the firms, at which point the firms can no longer directly increase reward.
-Similarly, every iteration firms' stored money is decreased by fixed operational costs as well as costs that scale with the size of the firm (i.e., $m \mapsto r(m - f)$, where $f$ is a fixed cost and $r$ is a multiplicative decay just below $1$).
+Similarly, every iteration firms' stored money is decreased by fixed operational costs.
+We added functionality to include costs that scale with the size of the firm (i.e., $m \mapsto r(m - f)$, where $f$ is a fixed cost and $r$ is a multiplicative decay just below $1$), but in practice these often led to unstable dynamics.
 Together, these additions allow for interesting model dynamics for the entirety of the experiment.
@@
 Therefore, convergence corresponds to the losses approaching zero.
-This occurs in our preliminary simulation, as can be seen in figure \ref{plt:losstozero}.
+This occurs in our preliminary simulation, as can be seen in Figure \ref{plt:losstozero}.
 
@@
 As such, we expect the price at which the firm sells goods to rise.
-In figure \ref{plt:firmprofitincrease}, we see that the firm's profit is generally increasing over time from an initial average profit of approximately $\$23$ per iteration to an average profit of up to $\$50$ per iteration.
+In Figure \ref{plt:firmprofitincrease}, we see that the firm's profit is generally increasing over time from an initial average profit of approximately $\$23$ per iteration to an average profit of up to $\$50$ per iteration.
 The noisiness in the increase in profit may be due to the complex, changing environment (due to the changing actions of the human agents), or the fact that the reward is actually the future-discounted logarithmic profit.
@@
 
+After verifying the preliminary results, we simulated a more complex scenario and tracking more standard economic metrics.
+We included five firm agents and 25 human agents using REINFORCE with episodes of length 200.
+The interest rate for human agents was set to $0.2\%$ and the operational cost for the firms was set to $\$10$.
+The money of each human agent for the first episode is shown in Figure \ref{plt:init-ppl}, and the money of each firm agent for the first episode is shown in Figure \ref{plt:init-firms}.
+The initial policies for human agents tended to buy goods greedily, while the initial policies for firm agents tended to sell greedily.
+
+\begin{figure}[h]
+  \begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+    \begin{axis}[
+      xlabel=Person Money (\$),
+      ylabel=Iteration
+    ]
+    \addplot table [y=mp0, x=i]{init.dat};
+    \addplot table [y=mp1, x=i]{init.dat};
+    \addplot table [y=mp2, x=i]{init.dat};
+    \addplot table [y=mp3, x=i]{init.dat};
+    \addplot table [y=mp4, x=i]{init.dat};
+    \addplot table [y=mp5, x=i]{init.dat};
+    \addplot table [y=mp6, x=i]{init.dat};
+    \addplot table [y=mp7, x=i]{init.dat};
+    \addplot table [y=mp8, x=i]{init.dat};
+    \addplot table [y=mp9, x=i]{init.dat};
+    \addplot table [y=mp10, x=i]{init.dat};
+    \addplot table [y=mp11, x=i]{init.dat};
+    \addplot table [y=mp12, x=i]{init.dat};
+    \addplot table [y=mp13, x=i]{init.dat};
+    \addplot table [y=mp14, x=i]{init.dat};
+    \addplot table [y=mp15, x=i]{init.dat};
+    \addplot table [y=mp16, x=i]{init.dat};
+    \addplot table [y=mp17, x=i]{init.dat};
+    \addplot table [y=mp18, x=i]{init.dat};
+    \addplot table [y=mp19, x=i]{init.dat};
+    \addplot table [y=mp20, x=i]{init.dat};
+    \addplot table [y=mp21, x=i]{init.dat};
+    \addplot table [y=mp22, x=i]{init.dat};
+    \addplot table [y=mp23, x=i]{init.dat};
+    \addplot table [y=mp24, x=i]{init.dat};
+    \end{axis}
+  \end{tikzpicture}
+  \caption{TODO.}
+  \label{plt:init-ppl}
+\end{figure}
+
+\begin{figure}[h]
+  \begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+    \begin{axis}[
+      xlabel=Firm Money (\$),
+      ylabel=Iteration
+    ]
+    \addplot table [y=mf0, x=i]{init.dat};
+    \addplot table [y=mf1, x=i]{init.dat};
+    \addplot table [y=mf2, x=i]{init.dat};
+    \addplot table [y=mf3, x=i]{init.dat};
+    \addplot table [y=mf4, x=i]{init.dat};
+    \end{axis}
+  \end{tikzpicture}
+  \caption{TODO.}
+  \label{plt:init-firms}
+\end{figure}
+
+At episode 60, distinct policies have emerged.
+The same plots of money for human and firm agents is shown in Figure \ref{plt:late-ppl} and Figure \ref{plt:late-firms}.
+In particular, a pattern that emerges amongst the human agent policies is the concept of what we call a \textit{retirement policy}.
+In the retirement policy, a human agent aims to increase its money for the majority of its life (similar to saving money for retirement).
+Between iterations 100 and 150, the policy adjusts to begin using the money to purchase goods and increase utility.
+In this case, the most successful retirement policy is executed by the TODO.
+It is also interesting to note that human agents that start out with more money
+
+\begin{figure}[h]
+  \begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+    \begin{axis}[
+      xlabel=Person Money (\$),
+      ylabel=Iteration
+    ]
+    \addplot table [y=mp0, x=i]{late.dat};
+    \addplot table [y=mp1, x=i]{late.dat};
+    \addplot table [y=mp2, x=i]{late.dat};
+    \addplot table [y=mp3, x=i]{late.dat};
+    \addplot table [y=mp4, x=i]{late.dat};
+    \addplot table [y=mp5, x=i]{late.dat};
+    \addplot table [y=mp6, x=i]{late.dat};
+    \addplot table [y=mp7, x=i]{late.dat};
+    \addplot table [y=mp8, x=i]{late.dat};
+    \addplot table [y=mp9, x=i]{late.dat};
+    \addplot table [y=mp10, x=i]{late.dat};
+    \addplot table [y=mp11, x=i]{late.dat};
+    \addplot table [y=mp12, x=i]{late.dat};
+    \addplot table [y=mp13, x=i]{late.dat};
+    \addplot table [y=mp14, x=i]{late.dat};
+    \addplot table [y=mp15, x=i]{late.dat};
+    \addplot table [y=mp16, x=i]{late.dat};
+    \addplot table [y=mp17, x=i]{late.dat};
+    \addplot table [y=mp18, x=i]{late.dat};
+    \addplot table [y=mp19, x=i]{late.dat};
+    \addplot table [y=mp20, x=i]{late.dat};
+    \addplot table [y=mp21, x=i]{late.dat};
+    \addplot table [y=mp22, x=i]{late.dat};
+    \addplot table [y=mp23, x=i]{late.dat};
+    \addplot table [y=mp24, x=i]{late.dat};
+    \end{axis}
+  \end{tikzpicture}
+  \caption{TODO.}
+  \label{plt:late-ppl}
+\end{figure}
+
+\begin{figure}[h]
+  \begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+    \begin{axis}[
+      xlabel=Firm Money (\$),
+      ylabel=Iteration
+    ]
+    \addplot table [y=mf0, x=i]{late.dat};
+    \addplot table [y=mf1, x=i]{late.dat};
+    \addplot table [y=mf2, x=i]{late.dat};
+    \addplot table [y=mf3, x=i]{late.dat};
+    \addplot table [y=mf4, x=i]{late.dat};
+    \end{axis}
+  \end{tikzpicture}
+  \caption{TODO.}
+  \label{plt:late-firms}
+\end{figure}
+
+\begin{figure}[h]
+  \begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+    \begin{axis}[
+      xlabel=Average Market Price of Goods,
+      ylabel=Iteration
+    ]
+    \addplot table [y=pg, x=i]{late.dat};
+    \end{axis}
+  \end{tikzpicture}
+  \caption{TODO.}
+  \label{plt:late-stockmarket}
+\end{figure}
+
+\begin{figure}[h]
+  \begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+    \begin{axis}[
+      xlabel=Total Goods Consumed,
+      ylabel=Iteration
+    ]
+    \addplot table [y=gp0, x=i]{late.dat};
+    \addplot table [y=gp1, x=i]{late.dat};
+    \addplot table [y=gp2, x=i]{late.dat};
+    \addplot table [y=gp3, x=i]{late.dat};
+    \addplot table [y=gp4, x=i]{late.dat};
+    \addplot table [y=gp5, x=i]{late.dat};
+    \addplot table [y=gp6, x=i]{late.dat};
+    \addplot table [y=gp7, x=i]{late.dat};
+    \addplot table [y=gp8, x=i]{late.dat};
+    \addplot table [y=gp9, x=i]{late.dat};
+    \addplot table [y=gp10, x=i]{late.dat};
+    \addplot table [y=gp11, x=i]{late.dat};
+    \addplot table [y=gp12, x=i]{late.dat};
+    \addplot table [y=gp13, x=i]{late.dat};
+    \addplot table [y=gp14, x=i]{late.dat};
+    \addplot table [y=gp15, x=i]{late.dat};
+    \addplot table [y=gp16, x=i]{late.dat};
+    \addplot table [y=gp17, x=i]{late.dat};
+    \addplot table [y=gp18, x=i]{late.dat};
+    \addplot table [y=gp19, x=i]{late.dat};
+    \addplot table [y=gp20, x=i]{late.dat};
+    \addplot table [y=gp21, x=i]{late.dat};
+    \addplot table [y=gp22, x=i]{late.dat};
+    \addplot table [y=gp23, x=i]{late.dat};
+    \addplot table [y=gp24, x=i]{late.dat};
+    \end{axis}
+  \end{tikzpicture}
+  \caption{TODO.}
+  \label{plt:late-goodspeople}
+\end{figure}
+
+\begin{figure}[h]
+  \begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+    \begin{axis}[
+      xlabel=GDP,
+      ylabel=Iteration
+    ]
+    \addplot table [y=gdp, x=i]{late.dat};
+    \end{axis}
+  \end{tikzpicture}
+  \caption{TODO.}
+  \label{plt:late-gdp}
+\end{figure}
+
+\begin{figure}[h]
+  \begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+    \begin{axis}[
+      xlabel=Gini Coefficient,
+      ylabel=Iteration
+    ]
+    \addplot table [y=gini, x=i]{late.dat};
+    \end{axis}
+  \end{tikzpicture}
+  \caption{TODO.}
+  \label{plt:late-gdp}
+\end{figure}
+
 \section{Conclusion}

### 2019-12-06 paper/paper.tex

@@
 This was chosen due to previous literature, and because such methods have stronger convergence guarantees than value-based reinforcement learning methods (though potentially to non-global optima).
+
+\medskip
+
+In policy gradient methods, instead of approximating the value functions and using a separate exploration strategy for choosing actions, we directly compute the policy.
+Given a policy model parameterized by $\theta$, we compute probabilities $\pi_{\theta}(a|s)$.
+When training this model, we aim to minimize the loss function \cite{actor-critic}:
+\[
+J(\theta) = -\sum_{s \in S} P_{\pi}(s) \sum_{a\in A} \pi_{\theta}(a|s) Q^{\pi}(s, a)
+\]
+Intuitively, a low loss means that the agent outputs high probabilities for high $Q$-valued actions.
+Policy Gradient Theorem allows us to express the gradient of this loss in a tractable manner \cite{reinforce-lectures} \cite{actor-critic}:
+\[
+\nabla_{\theta} J(\theta) = \mathbb{E}_{\pi}[Q^{\pi}(s, a) \nabla_{\theta} \log \pi_{\theta}(a|s)]
+\]
+This means that over a trajectory, we simply need to compute the gradients:
+\[
+\left(\sum_{i=t}^T \gamma^{i-t} r_t\right) \nabla_{\theta} \log \pi_{\theta}(a_t|s_t)
+\]
+We can then use standard gradient descent methods to optimize $\theta$.
 
@@
 
-Recall that the loss function for the REINFORCE algorithm is of the following form:
+Recall that the gradient the REINFORCE algorithm is of the following form:
 \[
-  \mathcal{L} = -\sum_{t = 1}^T \left[\log \pi_\theta(s_t, a_t) \left( \sum_{i = t}^n \gamma^{i - t}r_i \right) \right].  
+  \nabla_{\theta} J(\theta) = -\sum_{t = 1}^T \left[\nabla_{\theta} \log \pi_\theta(s_t, a_t) \left( \sum_{i = t}^n \gamma^{i - t}r_i \right) \right].  
 \]
 
-In particular, as the policy converges, the probability of taking a given action $a_t$ from state $s_t$ should approach one, and so the $\log \pi_\theta$ term will approach zero.
-Therefore, convergence corresponds to the losses approaching zero.
+In our training, we designate a proxy for the loss to reflect this gradient:
+\[
+  J'(\theta) = -\sum_{t = 1}^T \left[\log \pi_\theta(s_t, a_t) \left( \sum_{i = t}^n \gamma^{i - t}r_i \right) \right].  
+\]
+Note that, as the policy converges, the probability of taking a given action $a_t$ from state $s_t$ should approach one, and so the $\log \pi_\theta$ term will approach zero.
+Therefore, convergence corresponds to the (proxy) losses approaching zero.
 This occurs in our preliminary simulation, as can be seen in Figure \ref{plt:losstozero}.

### 2019-12-07 paper/paper.tex
stuff

@@
     \begin{axis}[
-      xlabel=Person Money (\$),
-      ylabel=Iteration
+      colormap={foo}{
+        % make a list of N colors
+        rgb255(1)=(0, 172, 168);
+        rgb255(2)=(0, 180, 164);
+        rgb255(3)=(0, 110, 199);
+        rgb255(4)=(0, 221, 144);
+        rgb255(5)=(0, 131, 189);
+        rgb255(6)=(0, 54, 227);
+        rgb255(7)=(0, 131, 189);
+        rgb255(8)=(0, 185, 162);
+        rgb255(9)=(0, 185, 162);
+        rgb255(10)=(0, 151, 179);
+        rgb255(11)=(0, 164, 172);
+        rgb255(12)=(0, 200, 154);
+        rgb255(13)=(0, 123, 193);
+        rgb255(14)=(0, 69, 220);
+        rgb255(15)=(0, 74, 217);
+        rgb255(16)=(0, 74, 217);
+        rgb255(17)=(0, 164, 172);
+        rgb255(18)=(0, 255, 127);
+        rgb255(19)=(0, 0, 255);
+        rgb255(20)=(0, 131, 189);
+        rgb255(21)=(0, 185, 162);
+        rgb255(22)=(0, 208, 150);
+        rgb255(23)=(0, 180, 164);
+        rgb255(24)=(0, 123, 193);
+        rgb255(25)=(0, 139, 185);
+      },
+      % use 0,...,N-1 
+      cycle list={[indices of colormap={0,...,24} of foo]},
+      xlabel=Iteration,
+      ylabel=Human Agent Money (\$)
     ]
@@
   \end{tikzpicture}
-  \caption{TODO.}
+  \caption{The money of each human agent during the first episode.
+All human agents appear to follow greedy policies, purchasing goods as quickly as possible before running out of money.
+The brighter colors correspond to higher skill levels.}
   \label{plt:init-ppl}
@@
     \begin{axis}[
-      xlabel=Firm Money (\$),
-      ylabel=Iteration
+      xlabel=Iteration,
+      ylabel=Firm Money (\$)
     ]
-    \addplot table [y=mf0, x=i]{init.dat};
-    \addplot table [y=mf1, x=i]{init.dat};
-    \addplot table [y=mf2, x=i]{init.dat};
-    \addplot table [y=mf3, x=i]{init.dat};
-    \addplot table [y=mf4, x=i]{init.dat};
+    \addplot +[mark=none] table [y=mf0, x=i]{init.dat};
+    \addplot +[mark=none] table [y=mf1, x=i]{init.dat};
+    \addplot +[mark=none] table [y=mf2, x=i]{init.dat};
+    \addplot +[mark=none] table [y=mf3, x=i]{init.dat};
+    \addplot +[mark=none] table [y=mf4, x=i]{init.dat};
     \end{axis}
   \end{tikzpicture}
-  \caption{TODO.}
+  \caption{The money of each firm agent during the first episode.
+All firms appear to follow policies that are complementary to the greedy policies present in the human agents.
+Once the human agents run out of money around iteration 100, the firms are no longer profitable.}
   \label{plt:init-firms}
@@
 In this case, the most successful retirement policy is executed by the TODO.
-It is also interesting to note that human agents that start out with more money
+We can see the total goods consumed by each human agent in Figure \ref{plt:late-goodspeople}, which shows that this retirement policy yields the maximum number of goods of any human agent over the entire episode.
+It is also interesting to note that human agents that start out with more money do not generally learn these policies.
 
@@
   \end{tikzpicture}
-  \caption{TODO.}
+  \caption{The money of each human agent during episode 60.
+The human agents have learned various policies, including a retirement policy in which the agent saves money and then spends money.
+The brighter colors correspond to higher skill levels.}
   \label{plt:late-ppl}
@@
     \begin{axis}[
-      xlabel=Firm Money (\$),
-      ylabel=Iteration
+      xlabel=Iteration,
+      ylabel=Firm Money (\$)
     ]
-    \addplot table [y=mf0, x=i]{late.dat};
-    \addplot table [y=mf1, x=i]{late.dat};
-    \addplot table [y=mf2, x=i]{late.dat};
-    \addplot table [y=mf3, x=i]{late.dat};
-    \addplot table [y=mf4, x=i]{late.dat};
+    \addplot +[mark=none] table [y=mf0, x=i]{late.dat};
+    \addplot +[mark=none] table [y=mf1, x=i]{late.dat};
+    \addplot +[mark=none] table [y=mf2, x=i]{late.dat};
+    \addplot +[mark=none] table [y=mf3, x=i]{late.dat};
+    \addplot +[mark=none] table [y=mf4, x=i]{late.dat};
     \end{axis}
   \end{tikzpicture}
-  \caption{TODO.}
+  \caption{The money of each firm agent during episode 60. Despite being identical, the diversity in policies led to some firms generating large profits, while others experienced less success.}
   \label{plt:late-firms}
@@
     \begin{axis}[
-      xlabel=Average Market Price of Goods,
-      ylabel=Iteration
-    ]
-    \addplot table [y=pg, x=i]{late.dat};
-    \end{axis}
-  \end{tikzpicture}
-  \caption{TODO.}
-  \label{plt:late-stockmarket}
-\end{figure}
-
-\begin{figure}[h]
-  \begin{tikzpicture}[every mark/.append style={mark size=1pt}]
-    \begin{axis}[
-      xlabel=Total Goods Consumed,
-      ylabel=Iteration
+      colormap={foo}{
+        % make a list of N colors
+        rgb255(1)=(0, 172, 168);
+        rgb255(2)=(0, 180, 164);
+        rgb255(3)=(0, 110, 199);
+        rgb255(4)=(0, 221, 144);
+        rgb255(5)=(0, 131, 189);
+        rgb255(6)=(0, 54, 227);
+        rgb255(7)=(0, 131, 189);
+        rgb255(8)=(0, 185, 162);
+        rgb255(9)=(0, 185, 162);
+        rgb255(10)=(0, 151, 179);
+        rgb255(11)=(0, 164, 172);
+        rgb255(12)=(0, 200, 154);
+        rgb255(13)=(0, 123, 193);
+        rgb255(14)=(0, 69, 220);
+        rgb255(15)=(0, 74, 217);
+        rgb255(16)=(0, 74, 217);
+        rgb255(17)=(0, 164, 172);
+        rgb255(18)=(0, 255, 127);
+        rgb255(19)=(0, 0, 255);
+        rgb255(20)=(0, 131, 189);
+        rgb255(21)=(0, 185, 162);
+        rgb255(22)=(0, 208, 150);
+        rgb255(23)=(0, 180, 164);
+        rgb255(24)=(0, 123, 193);
+        rgb255(25)=(0, 139, 185);
+      },
+      % use 0,...,N-1 
+      cycle list={[indices of colormap={0,...,24} of foo]},
+      xlabel=Iteration,
+      ylabel=Total Goods Consumed
     ]
@@
   \end{tikzpicture}
-  \caption{TODO.}
+  \caption{Total goods consumed by each human agent during episode 60.
+The most successful policy corresponds to the human agent in Figure \ref{plt:late-ppl} which ends with the most money (following a retirement policy).}
   \label{plt:late-goodspeople}
+\end{figure}
+
+We also investigate other relevant economic metrics in the model.
+For example, Figure \ref{plt:late-stockmarket} shows the average market price of a good for each iteration in the episode.
+Note that for this plots, as well as others, we have smoothed the results using a function adapted from the SciPy Cookbook \cite{scipy}.
+This curve is a rough estimate of the stock market in this economy.
+While the details of the curve are fairly chaotic, knowing the market price of a good is particularly useful because it allows us to estimate the total wealth of a given human agent at a given iteration.
+Wealth includes the current money of the human agent as well as the value of the goods owned by the human agent (as set by the current market price).
+From this, we compute the Gini coefficient for this population and plot it over time in Figure \ref{plt:late-gini}.
+We observe that inequality in this population tends ..
+
+\begin{figure}[h]
+  \begin{tikzpicture}[every mark/.append style={mark size=1pt}]
+    \begin{axis}[
+      xlabel=Iteration,
+      ylabel=Average Market Price of Goods
+    ]
+    \addplot [draw opacity=0.2] table [y=pg, x=i]{late.dat};
+    \addplot +[mark=none] table [y=ssm, x=i]{smoothed.dat};
+    \end{axis}
+  \end{tikzpicture}
+  \caption{The average market price of goods during episode 60. The transparent plot indicates the true values, and the superimposed red plot is a smoothed version.}
+  \label{plt:late-stockmarket}
 \end{figure}
@@
     \begin{axis}[
-      xlabel=GDP,
-      ylabel=Iteration
+      xlabel=Iteration,
+      ylabel=Gini Coefficient
     ]
-    \addplot table [y=gdp, x=i]{late.dat};
+    \addplot[draw opacity=0.2] table [y=gini, x=i]{late.dat};
+    \addplot +[mark=none] table [y=sgini, x=i]{smooth-gini.dat};
     \end{axis}
   \end{tikzpicture}
-  \caption{TODO.}
-  \label{plt:late-gdp}
+  \caption{The Gini coefficient during episode 60.
+The transparent plot indicates the true values, and the superimposed red plot is a smoothed version.
+The inequality in the population remains generally stable, but tends to trend upwards after 50 iterations.}
+  \label{plt:late-gini}
 \end{figure}
@@
     \begin{axis}[
-      xlabel=Gini Coefficient,
-      ylabel=Iteration
+      xlabel=Iteration,
+      ylabel=GDP
     ]
-    \addplot table [y=gini, x=i]{late.dat};
+    \addplot table [y=gdp, x=i]{late.dat};
     \end{axis}
   \end{tikzpicture}
-  \caption{TODO.}
+  \caption{The population GDP during episode 60. The transparent plot indicates true values, and the superimposed red plot is a smoothed version.}
   \label{plt:late-gdp}

### 2019-12-07 main.py

@@ -1,5 +1,6 @@
 import model
 import matplotlib.pyplot as plt
+from matplotlib import cm
 from constants import *
 from analytics import *
 from tqdm import tqdm
@@ -92,5 +93,37 @@ def main():
     reinforce_test()
     # plot1_for_paper()
 
+def read_file(name):
+    lines = []
+    with open(name, 'r') as f:
+        for line in f:
+            lines.append(line)
+
+    lines = [line.split() for line in lines]
+
+    pg_idx = lines[0].index('gini')
+    data = [float(x[pg_idx]) for x in lines[1:]]
+    data = smooth(data, 20)
+
+    with open('paper/smooth-gini.dat', 'w') as f:
+        for i in range(len(list(data))):
+            f.write(str(i) + ' ' + str(list(data)[i]) + '\n')
+    
+
+def thing():
+    colormap = cm.get_cmap('winter', 100)
+    skills = [1.06, 1.07, 0.97, 1.13, 1.0, 0.89, 1.0, 1.08, 1.08, 1.03, 1.05, 1.1, 0.99, 0.91, 0.92, 0.92, 1.05, 1.18, 0.81, 1.0, 1.08, 1.11, 1.07, 0.99, 1.01]
+    maxskill = max(skills)
+    minskill = min(skills)
+    dskill = maxskill - minskill
+    counter = 0
+    for skill in skills:
+        counter += 1
+        a = colormap((skill - minskill) / dskill)
+        a = (int(a[0] * 255), int(a[1] * 255), int(a[2] * 255))
+        print("rgb255(%s)=(%s, %s, %s);" % (counter, a[0], a[1], a[2]))
+
+
+
 if __name__ == "__main__":
-    main()
+    thing()

### 2019-12-07 paper/paper.tex
Final updates

@@
 In this work, we demonstrate how multi-agent reinforcement learning can be combined with a simple economic model to simulate and analyze macroeconomic phenomena under various conditions.
-In particular, we show how policy gradient learning algorithms yield qualitatively correct results in simple simulations, and we demonstrate nontrivial behavior in more complex scenarios involving over 30 separate agents.
+In particular, we show how policy gradient learning algorithms yield qualitatively correct results in simple simulations, and we demonstrate nontrivial behavior in more complex scenarios involving up to 30 separate agents.
 \end{abstract}
@@
 For more in-depth economic phenomena, one would likely benefit from including more markets and agents; for example, the five-sector economy model includes government, financial markets, and overseas influences.
-However, as we aim to provide a proof-of-concept for modeling economies with multi-agent reinforcement learning, the circular flow model serves as a simple foundation that can potentially produce nontrivial dynamics.
+However, as we aim to simply provide a proof-of-concept, the circular flow model is sufficient.
 
@@
 In the following work, we explore how various economic phenomena can be simulated and analyzed by applying reinforcement learning to the circular flow model.
-While we are limited by the simplicity of the model, there are many economic metrics that manifest in such a model.
+There are many economic metrics that manifest in such a model.
 For example, the gross domestic product (GDP) is widely touted as a comprehensive measure of a country's economic health.
@@
 Another important area of economics studies income and wealth inequality.
-A common approach to measuring these is to use the Gini coefficient.
-The Gini coefficient is a measure of the inequality of a distribution $p$ with mean $\mu$.
+A common approach to measuring these is to use the Gini coefficient: a measure of the inequality of a distribution $p$ with mean $\mu$.
 Explicitly, the Gini coefficient is given by
@@
 To address this, the author suggested applying an MLP or other generalizable model as an extension.
-The suggestion to combine policy gradient algorithms and a generalizable action-choosing model significanlty influenced our design.
+The suggestion to combine policy gradient algorithms and a generalizable action-choosing model significantly influenced our design.
 
@@
 
+\medskip
+In all subsequently discussed simulations, several parameter choices were fixed.
+The discount was set to $\gamma = 0.99$.
+The initial money given to firm agents was chosen by drawing a number for each firm from $\mathcal{U}(0, 1)$ and giving each firm money proportional to its number such that the total money given to firm agents was $\$10000N$.
+A similar process was repeated for the human agents except that the total money given to human agents was $\$1000M$.
+The human agent skill levels were drawn from $\mathcal{N}(1, 0.1)$.
+The firm operational cost was set to $\$10$, and the human agent interest rate was set to $(5/M)\%$.
+For all simulations, inverse demand curves were chosen over linear demand curves and, unless specified otherwise, the REINFORCE algorithm was used for learning.
+
 \subsection{Preliminary Results}
 
-To test the convergence of the learning and the ensure that the results were logical, we created a scenario with one firm agent and five human agents.
-The dynamics discussed below are for a simulation in which all agents utilized the REINFORCE algorithm, all agents used inverse demand curves, the human agent interest rate was set to $1\%$, the firm operational cost was set to $\$10$, and the discount was $\gamma = 0.99$.
-The firm was initialized with $\$10000$, while the human agents were initialized to have uniform random initial money such that the total money given to human agents was $\$5000$.
-The human skill levels (goods per hour) were drawn from $N(1, 0.1)$.
-The simulation was run for 100 iterations.
+To test the convergence of the learning and the ensure that the results were logical, we created a scenario with one firm agent and five human agents. The simulation was run for 100 iterations.
 
@@
 After verifying the preliminary results, we simulated a more complex scenario and tracking more standard economic metrics.
-We included five firm agents and 25 human agents using REINFORCE with episodes of length 200.
-The interest rate for human agents was set to $0.2\%$ and the operational cost for the firms was set to $\$10$.
+We included five firm agents and 25 human agents with episodes of length 200.
 The money of each human agent for the first episode is shown in Figure \ref{plt:init-ppl}, and the money of each firm agent for the first episode is shown in Figure \ref{plt:init-firms}.
@@
 
+\medskip
 At episode 60, distinct policies have emerged.
@@
 In particular, a pattern that emerges amongst the human agent policies is the concept of what we call a \textit{retirement policy}.
-In the retirement policy, a human agent aims to increase its money for the majority of its life (similar to saving money for retirement).
+In a retirement policy, a human agent aims to increase its money for the majority of its life (similar to saving money for retirement).
 Between iterations 100 and 150, the policy adjusts to begin using the money to purchase goods and increase utility.
-In this case, the most successful retirement policy is executed by the TODO.
-We can see the total goods consumed by each human agent in Figure \ref{plt:late-goodspeople}, which shows that this retirement policy yields the maximum number of goods of any human agent over the entire episode.
+In this case, the most successful retirement policy is executed by the green agent which has the most money of all agents at iteration 200.
+Further, we can see the total goods consumed by each human agent in Figure \ref{plt:late-goodspeople}, which shows that this retirement policy yields the maximum number of goods of any human agent over the entire episode.
 It is also interesting to note that human agents that start out with more money do not generally learn these policies.
@@
   \caption{Total goods consumed by each human agent during episode 60.
-The most successful policy corresponds to the human agent in Figure \ref{plt:late-ppl} which ends with the most money (following a retirement policy).}
+The most successful policy corresponds to the human agent in Figure \ref{plt:late-ppl}, which ends with the most money (following a retirement policy).}
   \label{plt:late-goodspeople}
@@
 
+\medskip
 We also investigate other relevant economic metrics in the model.
@@
     ]
-    \addplot table [y=gdp, x=i]{late.dat};
+    \addplot [draw opacity=0.2] table [y=gdp, x=i]{late.dat};
+    \addplot +[mark=none] table [y=sgdp, x=i]{smooth-gdp.dat};
     \end{axis}
   \end{tikzpicture}
-  \caption{The population GDP during episode 60. The transparent plot indicates true values, and the superimposed red plot is a smoothed version.}
+  \caption{The population GDP during episode 60.
+The transparent plot indicates true values, and the superimposed red plot is a smoothed version.
+The GDP trends downwards due to some human agent policies purchasing fewer goods after running out of money}
   \label{plt:late-gdp}

### 2019-12-07 paper/paper.tex

--- 2019-12-07 paper/paper.tex
@@
 
-After verifying the preliminary results, we simulated a more complex scenario and tracking more standard economic metrics.
-We included five firm agents and 25 human agents using REINFORCE with episodes of length 200.
+After verifying the preliminary results to evaluate the basic success of the RL algorithms, we began to analyze individual policies chosen by firms and people as well as the macroeconomic effects that resulted from multiple agents interacting in the environment under more complicated dynamics with government policies.
+
+Under the standard dynamics, human agents tend to converge on fairly trivial dynamics, immediately spending earned money on goods.
+Since the acquisition of goods ultimately drives reward, there is no incentive to save money, no matter the discount factor used.
+A particularly sharp demand curve for goods can impose a soft limit on a person’s buying rate, but otherwise does not change the strategy of spending early and exorbitantly.
+
+A side effect of this phenomenon is a pronounced market collapse, or recession.
+When a person runs out of money due to excessive spending, they are unable to buy any additional goods.
+The firms, in turn, note the reduced demand for goods and begin to produce fewer goods.
+Accordingly, they employ fewer people to produce goods.
+As a result, unemployment increases, and lesser-skilled people can no longer find work.
+Without a source of income, they spend their remaining money and drop out of the market.
+
+We found such recessions to be inescapable — the system can almost never recover, and the eventual equilibrium is one in which a single skilled person makes and buys all the goods -- a particularly unrealistic outcome.
+Since firms always up-sell goods (that is, sell them at a profit), it always requires fewer people to produce a set of goods than to consume them.
+The difference between these populations is thus unemployed and eventually unable to buy goods at all, and the employment rate steadily drops.
+People who lose all their money rarely interact in the market any further -- a situation eerily analogous to death.
+
+Interestingly, although this outcome is not optimal for most people and firms, it proved to be very stable, and most randomly-initialized systems converged to a recession.
+To rectify this behavior, we increased the interest rate, thereby incentives the saving of money, which can purchase more goods after the accumulation of interest.
+Under this policy, we find that people naturally develop \textit{retirement policies}, under which they spend the first half of their lives saving more than they spend, and the latter half spending more than they save.
+Although such policies maximize people’s utility, they do not appear to be easy to converge to.
+Although many people happily save money and continually spend the accrued interest, relatively few people learn to truly ``enjoy'' their retirement -- dramatically increasing spending near the end of their lives.
+
+As an example, we included five firm agents and 25 human agents using REINFORCE with episodes of length 200.
 The interest rate for human agents was set to $0.2\%$ and the operational cost for the firms was set to $\$10$.
 The money of each human agent for the first episode is shown in Figure \ref{plt:init-ppl}, and the money of each firm agent for the first episode is shown in Figure \ref{plt:init-firms}.
-The initial policies for human agents tended to buy goods greedily, while the initial policies for firm agents tended to sell greedily.
+As noted above, the initial policies for human agents tended to buy goods greedily, while the initial policies for firm agents tended to sell greedily.
 

### 2019-12-07 main.py
Commit to pull

@@ -101,11 +101,11 @@ def read_file(name):
 
     lines = [line.split() for line in lines]
 
-    pg_idx = lines[0].index('gini')
+    pg_idx = lines[0].index('gdp')
     data = [float(x[pg_idx]) for x in lines[1:]]
     data = smooth(data, 20)
 
-    with open('paper/smooth-gini.dat', 'w') as f:
+    with open('paper/smooth-gdp.dat', 'w') as f:
         for i in range(len(list(data))):
             f.write(str(i) + ' ' + str(list(data)[i]) + '\n')
     
@@ -126,4 +126,4 @@ def thing():
 
 
 if __name__ == "__main__":
-    thing()
+    read_file('paper/late.dat')

### 2019-12-07 paper/paper.tex

--- 2019-12-07 paper/paper.tex
@@
 
-After verifying the preliminary results, we simulated a more complex scenario and tracking more standard economic metrics.
-We included five firm agents and 25 human agents with episodes of length 200.
+After verifying the preliminary results to evaluate the basic success of the RL algorithms, we began to analyze individual policies chosen by firms and people as well as the macroeconomic effects that resulted from multiple agents interacting in the environment under more complicated dynamics with government policies.
+
+Under the standard dynamics, human agents tend to converge on fairly trivial dynamics, immediately spending earned money on goods.
+Since the acquisition of goods ultimately drives reward, there is no incentive to save money, no matter the discount factor used.
+A particularly sharp demand curve for goods can impose a soft limit on a person’s buying rate, but otherwise does not change the strategy of spending early and exorbitantly.
+
+A side effect of this phenomenon is a pronounced market collapse, or recession.
+When a person runs out of money due to excessive spending, they are unable to buy any additional goods.
+The firms, in turn, note the reduced demand for goods and begin to produce fewer goods.
+Accordingly, they employ fewer people to produce goods.
+As a result, unemployment increases, and lesser-skilled people can no longer find work.
+Without a source of income, they spend their remaining money and drop out of the market.
+
+We found such recessions to be inescapable — the system can almost never recover, and the eventual equilibrium is one in which a single skilled person makes and buys all the goods -- a particularly unrealistic outcome.
+Since firms always up-sell goods (that is, sell them at a profit), it always requires fewer people to produce a set of goods than to consume them.
+The difference between these populations is thus unemployed and eventually unable to buy goods at all, and the employment rate steadily drops.
+People who lose all their money rarely interact in the market any further -- a situation eerily analogous to death.
+
+Interestingly, although this outcome is not optimal for most people and firms, it proved to be very stable, and most randomly-initialized systems converged to a recession.
+To rectify this behavior, we increased the interest rate, thereby incentives the saving of money, which can purchase more goods after the accumulation of interest.
+Under this policy, we find that people naturally develop \textit{retirement policies}, under which they spend the first half of their lives saving more than they spend, and the latter half spending more than they save.
+Although such policies maximize people’s utility, they do not appear to be easy to converge to.
+Although many people happily save money and continually spend the accrued interest, relatively few people learn to truly ``enjoy'' their retirement -- dramatically increasing spending near the end of their lives.
+
+As an example, we included five firm agents and 25 human agents with episodes of length 200.
 The money of each human agent for the first episode is shown in Figure \ref{plt:init-ppl}, and the money of each firm agent for the first episode is shown in Figure \ref{plt:init-firms}.
-The initial policies for human agents tended to buy goods greedily, while the initial policies for firm agents tended to sell greedily.
+As noted above, the initial policies for human agents tended to buy goods greedily, while the initial policies for firm agents tended to sell greedily.
 
